Strict PPC thresholds were not fully met (@tbl-convergence-ppc-gate-appendix). Most flagged cells were Easy/VDT, with misfit concentrated in fast RT tails — a common issue for standard Wiener DDMs without fast-guess or across-trial variability parameters.

Mid-body RT quantiles and accuracy by condition matched the data reasonably well (@fig-ch3app-qp-plot, @fig-ch3app-ppc-rt-overlay). Effort and difficulty contrasts did not change much under alternative RT cutoffs (Appendix A4.2).

@fig-ch3app-ppc-rt-overlay compares observed and predicted RT densities. @fig-ch3app-qp-plot shows quantile–probability fit by condition.

```{r}
#| label: fig-ch3app-ppc-rt-overlay
#| fig-cap: "Posterior Predictive Check: RT Distributions. Empirical (black solid) vs. posterior predictive (blue solid) RT densities by Task × Effort × Difficulty. Overall model fit is good for central tendencies, with some misfit in fast tails (especially Easy/VDT)."
#| fig-align: center
#| out-width: "85%"
include_graphics_safe(fig_path("fig_ppc_rt_overlay.pdf"))
```

```{r}
#| label: fig-ch3app-qp-plot
#| fig-cap: "Quantile-Probability (QP) Plot. Empirical vs. predicted RT quantiles by difficulty level, with separate panels for Task × Effort. Points colored by difficulty (Standard=gray, Easy=blue, Hard=red) and shaped by response type (Correct/Error). Dashed diagonal = perfect prediction. Deviations primarily occur in fast tails for Easy/VDT conditions."
#| fig-align: center
#| out-width: "80%"
include_graphics_safe(fig_path("fig_qp.pdf"))
```
