This appendix provides detailed methods and diagnostic results for the posterior predictive checks (PPC) and model validation procedures reported in Chapter 3. Full PPC methods, primary PPC gate details, extended diagnostics, and parameter consistency checks are documented here.

## A3.1 PPC Validation Method {#sec-ch3-ppc-method}

Posterior Predictive Checks (PPC) were performed to validate model fit by comparing observed data to data simulated from the fitted model. To avoid aggregation bias (Jensen's Inequality) inherent in using group-level mean parameters in non-linear formulas, we used full posterior predictive sampling that respects subject-level random effects [@vehtari2020rank]. This approach generates predictions for every trial in the dataset, maintaining the hierarchical structure of the model.

**PPC Implementation**: For the primary model, we generated 1,000 posterior predictive draws using `brms::posterior_predict()` with `negative_rt = TRUE` to obtain signed reaction times (positive RT = "Different"/upper boundary, negative RT = "Same"/lower boundary). This parameter is critical for correctly extracting choice predictions from `brms` Wiener models. For each draw, we computed the proportion of "Different" responses and compared the distribution of predicted proportions to the observed proportion in the data.

**PPC Results**: On Standard trials, the model accurately predicted choice proportions: observed 10.9% "Different" responses vs. predicted 11.2% (95% credible interval: [9.9%, 12.7%]). The difference of 0.3% falls well within acceptable ranges, confirming that the model captures the data distribution accurately. The observed value falls within the 95% credible interval, indicating excellent model fit. Choice-proportion PPC validation is shown in @fig-ch3app-ppc-validation.

## A3.2 Primary PPC Gate: Subject-Wise Mid-Body Quantiles {#sec-ch3-ppc-gate}

Our **primary gate** for model acceptance is the subject-wise mid-body PPC (conditional on response, 2% censored). This metric respects individual differences and focuses on the core of the RT distribution, avoiding the Simpson's paradox issues inherent in pooled metrics and the known fast-tail limitations of the base Wiener DDM.

**Thresholds** (pre-declared):

-   QP RMSE fail \> 0.12 s (warn \> 0.09 s)
-   KS statistic fail \> 0.20 (warn \> 0.15)
-   Target: ≤ 15% of cells flagged

@tbl-ch3app-ppc-subj-cens shows subject-wise mid-body PPC results, and @tbl-ch3app-ppc-subj-summary shows the summary.

```{=latex}
\begin{landscape}
```

```{r}
#| label: tbl-ch3app-ppc-subj-cens
#| tbl-cap: "Subject-Wise Mid-Body PPC (30/50/70% quantiles; censored 2%)"
#| tbl-pos: "H"
if (!is.null(ppc_subj_cens)) {
  ppc_subj_cens %>%
    mutate(across(where(is.numeric), ~round(.x, 3))) %>%
    gt() %>%
    opt_table_font(size = "small") %>%
    tab_options(
      table.width = pct(90),
      table.font.size = px(9),
      column_labels.font.size = px(10),
      column_labels.padding = px(8),
      data_row.padding = px(6),
      table_body.hlines.width = px(0.5),
      table_body.vlines.width = px(0.5),
      table.border.top.width = px(1.5),
      table.border.bottom.width = px(1.5)
    )
} else {
  cat("Subject-wise PPC data not available.")
}
```

```{=latex}
\end{landscape}
```

```{r}
#| label: tbl-ch3app-ppc-subj-summary
#| tbl-cap: "Subject-Wise PPC Summary"
if (!is.null(pf_subj) && !is.na(pf_subj$n_cells)) {
  tibble(
    Metric = c("N Cells", "N Flagged", "% Flagged"),
    Value = c(
      as.character(pf_subj$n_cells),
      as.character(pf_subj$n_flagged),
      sprintf("%.1f%%", pf_subj$pct_flagged)
    )
  ) %>%
    gt()
}
```

**Result**: `r if (!is.null(pf_subj)) sprintf("%.1f%%", pf_subj$pct_flagged)` of cells flagged. The subject-wise mid-body PPC gate (based on strict pooled quantiles) was not met due to fast-tail deviations. However, as detailed below, the joint model cell-wise PPCs show that the model captures the central tendencies for the majority of conditions (Standard/Easy), with misfit primarily concentrated in VDT-Hard.

## A3.3 Visual PPC Diagnostics {#sec-ch3-ppc-visual}

RT overlay PPCs are shown in @fig-ch3app-ppc-rt-overlay. QP plot diagnostics are shown in @fig-ch3app-qp-plot. Cell-wise PPC residuals are shown in @fig-ch3app-pdiff-heatmap. Additional PPC diagnostics (best/median/worst cells) are shown in @fig-ch3app-ppc-small-multiples.

```{r}
#| label: fig-ch3app-pdiff-heatmap
#| fig-cap: "Observed vs. model-predicted p('different') across 12 cells (Task × Effort × Difficulty)."
#| fig-align: center
#| out-width: "90%"
include_graphics_safe(fig_path("fig_pdiff_heatmap.png"))
```

```{r}
#| label: fig-ch3app-ppc-small-multiples
#| fig-cap: ""
#| fig-align: center
#| out-width: "90%"
#| results: asis
if (knitr::is_latex_output()) {
  caption_text <- "PPC best/median/worst cells (QP RMSE and KS with thresholds)."
  caption_text_escaped <- latex_escape(caption_text)
  knitr::asis_output(paste0(
    "\\begin{landscape}\n",
    "\\begin{figure}[H]\n",
    "\\centering\n",
    "\\includegraphics[width=0.9\\textwidth]{", fig_path("fig_ppc_small_multiples.png"), "}\n",
    "\\caption{", caption_text_escaped, "}\n",
    "\\label{fig-ch3app-ppc-small-multiples}\n",
    "\\end{figure}\n",
    "\\end{landscape}\n"
  ))
} else {
  include_graphics_safe(fig_path("fig_ppc_small_multiples.png"))
  cat("\n\n*Figure: PPC best/median/worst cells (QP RMSE and KS with thresholds).*\n")
}
```

```{r}
#| label: fig-ch3app-ppc-validation
#| fig-cap: ""
#| fig-align: center
#| out-width: "90%"
#| results: asis
if (knitr::is_latex_output()) {
  caption_text <- "Posterior Predictive Check: Choice Proportions on Standard Trials. Histogram showing the distribution of predicted proportions of 'Different' responses across 500 posterior predictive draws. The red dashed vertical line indicates the observed proportion (10.9%), and the blue solid line indicates the mean predicted proportion (11.2%). The orange dotted lines show the 95% credible interval ([9.9%, 12.7%]). The observed value falls well within the predicted distribution, confirming that the hierarchical model accurately captures the data structure and avoids aggregation bias (Jensen's Inequality) by respecting subject-level heterogeneity."
  caption_text_escaped <- latex_escape(caption_text)
  knitr::asis_output(paste0(
    "\\begin{landscape}\n",
    "\\begin{figure}[H]\n",
    "\\centering\n",
    "\\includegraphics[width=0.9\\textwidth]{", fig_path("plot3_ppc_validation.png"), "}\n",
    "\\caption{", caption_text_escaped, "}\n",
    "\\label{fig-ch3app-ppc-validation}\n",
    "\\end{figure}\n",
    "\\end{landscape}\n"
  ))
} else {
  include_graphics_safe(fig_path("plot3_ppc_validation.png"))
  cat("\n\n*Figure: Posterior Predictive Check: Choice Proportions on Standard Trials. Histogram showing the distribution of predicted proportions of 'Different' responses across 500 posterior predictive draws. The red dashed vertical line indicates the observed proportion (10.9%), and the blue solid line indicates the mean predicted proportion (11.2%). The orange dotted lines show the 95% credible interval ([9.9%, 12.7%]). The observed value falls well within the predicted distribution, confirming that the hierarchical model accurately captures the data structure and avoids aggregation bias (Jensen's Inequality) by respecting subject-level heterogeneity.*\n")
}
```

```{r}
#| label: fig-ch3app-ppc-rt-overlay
#| fig-cap: "Posterior Predictive Check: RT Distributions. Empirical (black solid) vs. posterior predictive (blue solid) RT densities by Task × Effort × Difficulty. Overall model fit is good for central tendencies, with some misfit in fast tails (especially Easy/VDT)."
#| fig-align: center
#| out-width: "90%"
include_graphics_safe(fig_path("fig_ppc_rt_overlay.pdf"))
```

```{r}
#| label: fig-ch3app-qp-plot
#| fig-cap: "Quantile-Probability (QP) Plot. Empirical vs. predicted RT quantiles by difficulty level, with separate panels for Task × Effort. Points colored by difficulty (Standard=gray, Easy=blue, Hard=red) and shaped by response type (Correct/Error). Dashed diagonal = perfect prediction. Deviations primarily occur in fast tails for Easy/VDT conditions."
#| fig-align: center
#| out-width: "90%"
include_graphics_safe(fig_path("fig_qp.pdf"))
```

### A3.3.1 Sensitivity Analyses {#sec-ch3-ppc-sensitivity}

We conducted additional sensitivity analyses (Unconditional Pooled PPC, Conditional Pooled PPC) which confirmed that the core findings are robust, though strict pooled metrics flag more cells due to fast-tail misfit. These additional checks are detailed in the Supplementary Figures.

## A3.5 Parameter Consistency and Sanity Checks {#sec-ch3-ppc-sanity}

To validate the internal consistency of our model estimates, we performed three sanity checks recommended by independent expert review. These checks verify that parameter estimates are mathematically consistent with observed behavioral patterns. All sanity-check figures are provided in @fig-ch3app-sanity-rt-asymmetry–@fig-ch3app-sanity-subject-heterogeneity.

### RT Asymmetry on Standard Trials

On Standard trials, the model estimated a strong negative drift rate (v = -1.404) toward "Same" responses, combined with a slight starting-point bias toward "Different" (z = 0.567). To verify the consistency of these estimates, we examined whether RT patterns align with model predictions.

**Result**: "Same" responses were significantly faster than "Different" responses (mean RT: 1.03 s vs. 1.32 s, difference = 293 ms). This pattern aligns perfectly with the model's prediction: strong negative drift causes rapid evidence accumulation toward "Same", resulting in fast "Same" responses. The slower "Different" responses likely reflect rare errors that occur when the process fails to reach the "Same" boundary within the response window.

```{r}
#| label: fig-ch3app-sanity-rt-asymmetry
#| fig-cap: "RT Distribution by Response Type on Standard Trials. Density plots showing reaction time distributions for \"Same\" (crimson) and \"Different\" (blue) responses on Standard (Δ=0) trials. \"Same\" responses were 293 ms faster on average (mean RT = 1.03 s) than \"Different\" responses (mean RT = 1.32 s). This pattern is consistent with strong negative drift: when participants rapidly accumulate evidence for identity, they reach the \"Same\" boundary quickly, resulting in fast RTs. \"Different\" responses (errors) occur less frequently and typically require more time, possibly reflecting near-deadline responses when the process fails to reach the \"Same\" boundary before the response window closes."
#| fig-align: center
#| out-width: "90%"
include_graphics_safe(fig_path("sanity_check1_rt_asymmetry.png"))
```

### Hard Trial Drift Direction

The primary model estimated that Hard trials have negative drift relative to Standard trials. To verify this estimate is consistent with observed below-chance accuracy on Hard trials, we examined the posterior distribution of Hard trial drift rates.

**Result**: Hard trials show consistently negative drift (mean v = -0.643, 95% CrI: [-0.740, -0.546], P(v < 0) = 100%). This confirms that the sensory evidence for difference on Hard trials is too weak to overcome the baseline tendency toward "Same", explaining the observed below-chance accuracy (~30%) on Hard trials.

```{r}
#| label: fig-ch3app-sanity-hard-drift
#| fig-cap: "Posterior Distribution of Hard Trial Drift Rate. Density plot showing the posterior distribution of drift rate (v) for Hard trials. The mean drift rate is -0.643 (95% CrI: [-0.740, -0.546]), with 100% of posterior draws below zero. This confirms that Hard trials have negative drift toward \"Same\" responses, explaining why participants choose \"Same\" approximately 70% of the time even when stimuli differ. The weak sensory evidence for difference (small stimulus offsets) is insufficient to overcome the baseline tendency to see stimuli as identical."
#| fig-align: center
#| out-width: "90%"
include_graphics_safe(fig_path("sanity_check2_hard_drift.png"))
```

### Subject Heterogeneity in Drift Rates

The discrepancy between analytical predictions (using group-level mean parameters) and PPC results (using full posterior with subject heterogeneity) suggests substantial individual differences in drift rates. To verify this, we examined the distribution of subject-level drift rate estimates.

**Result**: Subject-level drift rates show substantial heterogeneity (SD = 0.65, range: -3.08 to -0.21). Most subjects (60%) show strong negative drift (|v| ≥ 1.0), while a small subset (4.5%) show weak drift (|v| < 0.5). This heterogeneity explains the PPC vs. analytical formula discrepancy: subjects with weaker drift contribute disproportionately to error rates, but their contribution is masked when using group-level mean parameters.

**Conclusion**: All three sanity checks confirm the internal consistency of our model estimates. RT patterns, drift directions, and individual differences align with model predictions, providing strong evidence that the hierarchical DDM accurately captures the decision-making processes in our data.

```{r}
#| label: fig-ch3app-sanity-subject-heterogeneity
#| fig-cap: "Distribution of Subject-Level Drift Rates on Standard Trials. Histogram showing the distribution of subject-specific drift rate estimates (deviations from group mean) for Standard trials. The distribution shows substantial heterogeneity (SD = 0.65, range: -3.08 to -0.21), with most subjects showing strong negative drift (|v| ≥ 1.0) but a small subset (4.5%) showing weak drift (|v| < 0.5). This heterogeneity explains why the analytical formula (using group means) under-predicts error rates compared to PPC (which respects individual differences), consistent with Jensen's Inequality in hierarchical models."
#| fig-align: center
#| out-width: "90%"
include_graphics_safe(fig_path("sanity_check3_subject_heterogeneity.png"))
```

## A3.6 Additional Diagnostic Figures and Tables

```{r}
#| label: tbl-ch3app-ppc-gate
#| tbl-cap: "Pooled PPC Gate Summary (Strict Test)"
if (!is.null(ppc_gate)) {
  ppc_gate %>%
    mutate(across(where(is.numeric), ~round(.x, 3))) %>%
    gt() %>%
    cols_label(
      n_cells = "N Cells",
      pct_flagged = "% Flagged",
      max_qp = "Max QP RMSE",
      max_ks = "Max KS"
    )
}
```

Additional diagnostic figures and tables are provided in the Supplementary Figures section of the main chapter.
