---
title: "Pupil Data Quality Report: Data Availability and Analysis Readiness for Chapters 2 & 3"
subtitle: "BAP Study - Pupillometry Analysis Pipeline"
author: "Mohammad Dastgheib"
date: now
date-format: "MMMM D, YYYY [at] hh:mm A"
bibliography: references.bib
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-tools: true
    theme: flatly
    fig-width: 10
    fig-height: 6
    df-print: paged
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(tidyr)
  library(ggplot2)
  library(plotly)
  library(knitr)
  library(kableExtra)
  library(scales)
  library(patchwork)
})

# Determine paths
REPO_ROOT <- if (file.exists("02_pupillometry_analysis")) {
  normalizePath(getwd())
} else if (file.exists("../02_pupillometry_analysis")) {
  normalizePath("..")
} else {
  normalizePath(getwd())
}

# Use processed pupil data (most recent version with fixed baseline alignment)
# Location: data/pupil_processed/ (formerly quick_share_v7)
pupil_data_dir <- file.path(REPO_ROOT, "data", "pupil_processed")
pupil_qc_dir <- file.path(pupil_data_dir, "qc")
pupil_analysis_dir <- file.path(pupil_data_dir, "analysis_ready")

if (!dir.exists(pupil_data_dir)) {
  stop("data/pupil_processed directory not found. Please ensure the data processing pipeline has been run.")
}

# Load QC data from processed pupil data
gate_rates <- read_csv(file.path(pupil_qc_dir, "02_gate_pass_rates_by_task_threshold.csv"), show_col_types = FALSE)
join_health <- read_csv(file.path(pupil_qc_dir, "01_join_health_by_subject_task.csv"), show_col_types = FALSE)
auc_rates <- read_csv(file.path(pupil_qc_dir, "08_auc_non_na_rates.csv"), show_col_types = FALSE)
auc_missing <- read_csv(file.path(pupil_qc_dir, "03_auc_missingness_reasons.csv"), show_col_types = FALSE)

# Load analysis-ready datasets
ch2_data <- read_csv(file.path(pupil_analysis_dir, "ch2_triallevel.csv"), show_col_types = FALSE)
ch3_data <- read_csv(file.path(pupil_analysis_dir, "ch3_triallevel.csv"), show_col_types = FALSE)

# Load merged data for additional statistics
merged_file <- file.path(pupil_data_dir, "merged", "BAP_triallevel_merged_v4.csv")
if (file.exists(merged_file)) {
  merged_data <- read_csv(merged_file, show_col_types = FALSE)
} else {
  warning("Merged data file not found, using analysis-ready files only")
  merged_data <- NULL
}
```

## Executive Summary

### What I Need From You

::: {.callout-note}
### Decision Points

Please review and approve the following:

1. **Quality Thresholds**: Review current default thresholds for Chapter 2 (60%) and Chapter 3 (50% + RT filter). Note: Final threshold selection will be determined based on sensitivity analyses in dedicated analysis scripts.
2. **AUC Requirement for Pupil-Enhanced DDM**: Approve whether to require `auc_available == TRUE` for the pupil-predictor DDM subset (see Section 6.2 for intersection counts)
3. **Sensitivity Analysis**: Approve whether to run and report 50% and 70% threshold sensitivity analyses as robustness checks (even if brief)
:::

---

This report provides a comprehensive overview of the pupil data
collected in the BAP (Brain, Aging and Perception) study, with a focus
on data quality, availability, and readiness for two dissertation
chapters. **This report establishes the foundation (data availability, retention rates, quality metrics) but final threshold recommendations will be determined based on sensitivity analyses conducted in dedicated Chapter 2 and Chapter 3 analysis scripts**, comparing effect sizes, standard errors, and statistical significance across threshold combinations:

-   **Chapter 2**: Psychometric sensitivity and pupil-indexed arousal
    coupling analyses
-   **Chapter 3**: Drift Diffusion Model (DDM) analyses with pupil
    predictors

```{r exec-summary}
# Calculate key metrics from quick_share_v7 data
n_subjects <- length(unique(ch2_data$sub))
total_trials <- nrow(ch2_data)

# Behavioral join coverage
if (!is.null(merged_data)) {
  total_behavioral <- sum(merged_data$has_behavioral_data, na.rm = TRUE)
  behavioral_rate <- 100 * total_behavioral / nrow(merged_data)
} else {
  # Estimate from join_health
  total_behavioral <- sum(join_health$n_behavioral, na.rm = TRUE)
  behavioral_rate <- 100 * sum(join_health$n_behavioral, na.rm = TRUE) / sum(join_health$n_total, na.rm = TRUE)
}

# Chapter 2 metrics (using gate_pupil_primary flag if available, otherwise use threshold)
if ("gate_pupil_primary" %in% names(ch2_data)) {
  ch2_primary_total <- sum(ch2_data$gate_pupil_primary, na.rm = TRUE)
} else if ("pass_primary_060" %in% names(ch2_data)) {
  ch2_primary_total <- sum(ch2_data$pass_primary_060, na.rm = TRUE)
} else {
  # Use gate rates at 0.60 threshold
  ch2_primary_total <- sum(gate_rates$n_pass_060, na.rm = TRUE)
}
ch2_rate <- ifelse(total_trials > 0, 100 * ch2_primary_total / total_trials, 0)

# Chapter 3 metrics (using ddm_ready flag if available)
if ("ddm_ready" %in% names(ch3_data)) {
  ch3_ready_total <- sum(ch3_data$ddm_ready, na.rm = TRUE)
} else {
  # Estimate from gate rates at 0.50 threshold
  ch3_ready_total <- sum(gate_rates$n_pass_050, na.rm = TRUE)
}
ch3_rate <- ifelse(nrow(ch3_data) > 0, 100 * ch3_ready_total / nrow(ch3_data), 0)

# AUC availability
auc_available_both <- auc_rates %>% filter(task == "ALL") %>% pull(auc_available_both_pct)
auc_available_adt <- auc_rates %>% filter(task == "ADT") %>% pull(auc_available_both_pct)
auc_available_vdt <- auc_rates %>% filter(task == "VDT") %>% pull(auc_available_both_pct)

# Task breakdown from gate rates
task_summary <- gate_rates %>%
  select(task, n_total, n_pass_050, n_pass_060, n_pass_070, n_auc_ready) %>%
  mutate(
    pct_pass_050 = 100 * n_pass_050 / n_total,
    pct_pass_060 = 100 * n_pass_060 / n_total,
    pct_auc = 100 * n_auc_ready / n_total
  )

# Calculate Chapter 2 and Chapter 3 ready trials by task
task_summary_ch2 <- ch2_data %>%
  group_by(task) %>%
  summarise(
    n_ch2_ready = if ("gate_pupil_primary" %in% names(ch2_data)) {
      sum(gate_pupil_primary, na.rm = TRUE)
    } else if ("pass_primary_060" %in% names(ch2_data)) {
      sum(pass_primary_060, na.rm = TRUE)
    } else {
      sum(baseline_quality >= 0.60 & cog_quality >= 0.60, na.rm = TRUE)
    },
    .groups = "drop"
  )

task_summary_ch3 <- ch3_data %>%
  group_by(task) %>%
  summarise(
    n_ch3_ready = if ("ddm_ready" %in% names(ch3_data)) {
      sum(ddm_ready, na.rm = TRUE)
    } else {
      # Fallback: use B1_quality if available, otherwise baseline_quality
      if ("B1_quality" %in% names(ch3_data)) {
        sum(B1_quality >= 0.50 & cog_quality >= 0.50 & 
            !is.na(rt) & rt >= 0.2 & rt <= 3.0, na.rm = TRUE)
      } else {
        sum(baseline_quality >= 0.50 & cog_quality >= 0.50 & 
            !is.na(rt) & rt >= 0.2 & rt <= 3.0, na.rm = TRUE)
      }
    },
    .groups = "drop"
  )

# Merge with task_summary
task_summary <- task_summary %>%
  left_join(task_summary_ch2, by = "task") %>%
  left_join(task_summary_ch3, by = "task") %>%
  mutate(
    pct_ch2 = ifelse(n_total > 0, 100 * n_ch2_ready / n_total, 0),
    pct_ch3 = ifelse(n_total > 0, 100 * n_ch3_ready / n_total, 0),
    across(where(is.numeric), ~ if_else(is.na(.x), 0, .x))
  )
```

### Key Findings

-   **Total Participants**: `r n_subjects` participants with pupil data
-   **Total Trials Available**:
    -   Total trials in dataset:
        `r format(total_trials, big.mark = ",")`
    -   Trials with behavioral data:
        `r format(total_behavioral, big.mark = ",")`
        (`r sprintf("%.1f", behavioral_rate)`% behavioral join rate)
    -   Trials with both AUC metrics:
        `r sprintf("%.1f", auc_available_both)`% overall
-   **Chapter 2 Readiness** (Primary threshold: 60% validity):
    -   Usable trials: `r format(ch2_primary_total, big.mark = ",")`
        (`r sprintf("%.1f", ch2_rate)`% of total trials)
    -   Analysis-ready dataset:
        `r format(nrow(ch2_data), big.mark = ",")` trials in
        `ch2_triallevel.csv` (contains ALL pupil trials; filter by `gate_pupil_primary == TRUE` to get the `r format(ch2_primary_total, big.mark = ",")` analysis-ready trials)
-   **Chapter 3 Readiness** (Primary threshold: 50% validity +
    behavioral RT filter):
    -   Pupil+behavior-ready trials:
        `r format(ch3_ready_total, big.mark = ",")`
        (`r sprintf("%.1f", ch3_rate)`% of total trials)
    -   Analysis-ready dataset:
        `r format(nrow(ch3_data), big.mark = ",")` trials in
        `ch3_triallevel.csv` (contains ALL pupil trials; filter by `ddm_ready == TRUE` to get the `r format(ch3_ready_total, big.mark = ",")` analysis-ready trials)

*Note: Both analysis-ready datasets contain the same
`r format(nrow(ch2_data), big.mark = ",")` trials (ALL pupil trials from the merged dataset). They differ only in
the quality flags provided (`gate_pupil_primary` for Chapter 2,
`ddm_ready` for Chapter 3), allowing us to filter to the trials that
meet each chapter's quality criteria. The datasets include trials both WITH and WITHOUT behavioral data; filter by the quality flags to get the analysis-ready subsets.*

-   **AUC Availability** (Both Total AUC and Cognitive AUC):
    -   Overall: `r sprintf("%.1f", auc_available_both)`%
    -   ADT: `r sprintf("%.1f", auc_available_adt)`%
    -   VDT: `r sprintf("%.1f", auc_available_vdt)`%

### Data by Task

```{r task-summary-table}
task_display <- task_summary %>%
  select(task, n_total, n_ch2_ready, pct_ch2, n_ch3_ready, pct_ch3, n_auc_ready, pct_auc) %>%
  mutate(across(where(is.numeric), ~ if_else(is.na(.x), 0, .x)))

kable(task_display,
      col.names = c("Task", "Total Trials", "Ch2 Ready (60%)", 
                    "Ch2 %", "Ch3 Ready (50%+RT)", "Ch3 %", "AUC Ready", "AUC %"),
      digits = c(0, 0, 0, 1, 0, 1, 0, 1),
      caption = "Data Availability by Task (Chapters 2 & 3)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## 2. Introduction and Research Context

### 2.1 Study Overview

The BAP study examines how physical effort and cognitive demands
interact in older adults using a dual-task paradigm combining handgrip
force manipulation with perceptual discrimination tasks. Pupillometry
provides a continuous, non-invasive measure of arousal that reflects
both physical effort (tonic arousal) and cognitive engagement (phasic
arousal).

### 2.2 Data Collection

-   **Tasks**: Auditory Discrimination Task (ADT) and Visual
    Discrimination Task (VDT)
-   **Sessions**: Scanner sessions 2-3 (session 1/practice excluded)
-   **Runs**: Up to 5 runs per task per session
-   **Sampling Rate**: 250 Hz (4 ms per sample)
-   **Trial Structure**: Each trial includes baseline, squeeze, stimulus
    presentation, and response periods

### 2.3 Dissertation Chapters

**Chapter 2: Psychometric-Pupil Coupling**

-   Tests how trial-wise pupil-indexed arousal (Cognitive AUC) modulates
    psychometric sensitivity
-   Requires high-quality baseline and cognitive window data (≥60%
    validity for primary analyses)
-   Uses Total AUC for effort manipulation checks and Cognitive AUC for
    psychometric coupling
-   **Quality gates**: `baseline_quality >= 0.60` AND
    `cog_quality >= 0.60` (from MATLAB pipeline quality metrics)

**Chapter 3: DDM with Pupil Predictors**

-   Examines how pupil-indexed arousal influences decision-making
    processes (drift rate, boundary separation)
-   Requires behavioral RT data (0.2-3.0s) plus moderate-quality pupil
    data (≥50% validity)
-   Can also run behavior-only DDM models without pupil requirements

## 3. Overall Data Availability

### 3.1 Participant-Level Summary

```{r participant-summary}
# Chapter 2 participant summary
participant_summary_ch2 <- ch2_data %>%
  group_by(sub) %>%
  summarise(
    n_tasks = n_distinct(task),
    n_sessions = n_distinct(session_used),
    total_trials_ch2 = n(),
    total_trials_ch2_ready = if ("gate_pupil_primary" %in% names(ch2_data)) {
      sum(gate_pupil_primary, na.rm = TRUE)
    } else if ("pass_primary_060" %in% names(ch2_data)) {
      sum(pass_primary_060, na.rm = TRUE)
    } else {
      # Fallback: use B1_quality if available, otherwise baseline_quality
      if ("B1_quality" %in% names(ch2_data)) {
        sum(B1_quality >= 0.60 & cog_quality >= 0.60, na.rm = TRUE)
      } else {
        sum(baseline_quality >= 0.60 & cog_quality >= 0.60, na.rm = TRUE)
      }
    },
    total_trials_auc_ch2 = sum(auc_available == TRUE, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    ch2_rate = ifelse(total_trials_ch2 > 0, 100 * total_trials_ch2_ready / total_trials_ch2, 0)
  )

# Chapter 3 participant summary
participant_summary_ch3 <- ch3_data %>%
  group_by(sub) %>%
  summarise(
    total_trials_ch3 = n(),
    total_trials_ch3_ready = if ("ddm_ready" %in% names(ch3_data)) {
      sum(ddm_ready, na.rm = TRUE)
    } else {
      # Fallback: use B1_quality if available, otherwise baseline_quality
      if ("B1_quality" %in% names(ch3_data)) {
        sum(B1_quality >= 0.50 & cog_quality >= 0.50 & 
            !is.na(rt) & rt >= 0.2 & rt <= 3.0, na.rm = TRUE)
      } else {
        sum(baseline_quality >= 0.50 & cog_quality >= 0.50 & 
            !is.na(rt) & rt >= 0.2 & rt <= 3.0, na.rm = TRUE)
      }
    },
    total_trials_auc_ch3 = sum(auc_available == TRUE, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    ch3_rate = ifelse(total_trials_ch3 > 0, 100 * total_trials_ch3_ready / total_trials_ch3, 0)
  )

# Combine summaries
participant_summary <- participant_summary_ch2 %>%
  full_join(participant_summary_ch3, by = "sub") %>%
  mutate(
    total_trials_ch2 = ifelse(is.na(total_trials_ch2), 0, total_trials_ch2),
    total_trials_ch2_ready = ifelse(is.na(total_trials_ch2_ready), 0, total_trials_ch2_ready),
    total_trials_ch3 = ifelse(is.na(total_trials_ch3), 0, total_trials_ch3),
    total_trials_ch3_ready = ifelse(is.na(total_trials_ch3_ready), 0, total_trials_ch3_ready),
    ch2_rate = ifelse(is.na(ch2_rate), 0, ch2_rate),
    ch3_rate = ifelse(is.na(ch3_rate), 0, ch3_rate)
  )

# Summary statistics
participant_stats <- participant_summary %>%
  summarise(
    median_trials_ch2 = median(total_trials_ch2, na.rm = TRUE),
    q25_trials_ch2 = quantile(total_trials_ch2, 0.25, na.rm = TRUE),
    q75_trials_ch2 = quantile(total_trials_ch2, 0.75, na.rm = TRUE),
    median_trials_ch3 = median(total_trials_ch3, na.rm = TRUE),
    q25_trials_ch3 = quantile(total_trials_ch3, 0.25, na.rm = TRUE),
    q75_trials_ch3 = quantile(total_trials_ch3, 0.75, na.rm = TRUE),
    median_trials_ch2_ready = median(total_trials_ch2_ready, na.rm = TRUE),
    median_trials_ch3_ready = median(total_trials_ch3_ready, na.rm = TRUE),
    median_ch2_rate = median(ch2_rate, na.rm = TRUE),
    median_ch3_rate = median(ch3_rate, na.rm = TRUE)
  )

cat("**Per-Participant Summary Statistics:**\n\n")
cat("**Total Trials (Both Chapters):**\n")
cat("- Median total trials per participant: ", participant_stats$median_trials_ch2, 
    " (IQR: ", participant_stats$q25_trials_ch2, "-", participant_stats$q75_trials_ch2, ")\n", sep = "")
cat("*Note: Both Chapter 2 and Chapter 3 datasets contain the same total number of trials per participant (all pupil trials from the merged dataset).*\n\n", sep = "")

cat("**Chapter 2 (60% Quality Threshold):**\n")
cat("- Median Chapter 2 ready trials per participant: ", participant_stats$median_trials_ch2_ready, "\n", sep = "")
cat("- Median Chapter 2 retention rate per participant: ", sprintf("%.1f", participant_stats$median_ch2_rate), "%\n\n", sep = "")

cat("**Chapter 3 (50% Quality Threshold + RT Filter):**\n")
cat("- Median Chapter 3 ready trials per participant: ", participant_stats$median_trials_ch3_ready, "\n", sep = "")
cat("- Median Chapter 3 retention rate per participant: ", sprintf("%.1f", participant_stats$median_ch3_rate), "%\n\n", sep = "")

cat("*Note: Chapter 3 has higher retention rates because it uses a more lenient quality threshold (50% vs 60%) and only requires behavioral RT data (0.2-3.0s). This prioritizes sample size for DDM analyses, while Chapter 2 prioritizes high-quality data for psychometric coupling analyses.*\n", sep = "")
```

```{r participant-distribution-ch2-trials, fig.height=5}
p1_data <- participant_summary %>%
  filter(total_trials_ch2_ready > 0)

p1 <- p1_data %>%
  ggplot(aes(x = total_trials_ch2_ready, text = paste0("Subject: ", sub, "<br>",
                                                         "Ch2 Ready Trials: ", total_trials_ch2_ready, "<br>",
                                                         "Total Trials: ", total_trials_ch2, "<br>",
                                                         "Retention Rate: ", round(ch2_rate, 1), "%"))) +
  geom_histogram(bins = 30, fill = "darkgreen", alpha = 0.7, color = "white") +
  labs(
    x = "Chapter 2 Ready Trials",
    y = "Number of Participants",
    title = "Distribution of Chapter 2 Ready Trials Across Participants"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"))

p1_ly <- ggplotly(p1, tooltip = "text")
print(p1_ly)
```

```{r participant-distribution-ch3-trials, fig.height=5}
p2_data <- participant_summary %>%
  filter(total_trials_ch3_ready > 0)

p2 <- p2_data %>%
  ggplot(aes(x = total_trials_ch3_ready, text = paste0("Subject: ", sub, "<br>",
                                                         "Ch3 Ready Trials: ", total_trials_ch3_ready, "<br>",
                                                         "Total Trials: ", total_trials_ch3, "<br>",
                                                         "Retention Rate: ", round(ch3_rate, 1), "%"))) +
  geom_histogram(bins = 30, fill = "darkblue", alpha = 0.7, color = "white") +
  labs(
    x = "Chapter 3 Ready Trials",
    y = "Number of Participants",
    title = "Distribution of Chapter 3 Ready Trials Across Participants"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"))

p2_ly <- ggplotly(p2, tooltip = "text")
print(p2_ly)
```

```{r participant-distribution-ch2-rate, fig.height=5}
p3_data <- participant_summary %>%
  filter(total_trials_ch2 > 0)

p3 <- p3_data %>%
  ggplot(aes(x = ch2_rate, text = paste0("Subject: ", sub, "<br>",
                                          "Retention Rate: ", round(ch2_rate, 1), "%", "<br>",
                                          "Ch2 Ready Trials: ", total_trials_ch2_ready, "<br>",
                                          "Total Trials: ", total_trials_ch2))) +
  geom_histogram(bins = 30, fill = "darkgreen", alpha = 0.7, color = "white") +
  labs(
    x = "Chapter 2 Retention Rate (%)",
    y = "Number of Participants",
    title = "Distribution of Chapter 2 Retention Rates"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"))

p3_ly <- ggplotly(p3, tooltip = "text")
print(p3_ly)
```

```{r participant-distribution-ch3-rate, fig.height=5}
p4_data <- participant_summary %>%
  filter(total_trials_ch3 > 0)

p4 <- p4_data %>%
  ggplot(aes(x = ch3_rate, text = paste0("Subject: ", sub, "<br>",
                                          "Retention Rate: ", round(ch3_rate, 1), "%", "<br>",
                                          "Ch3 Ready Trials: ", total_trials_ch3_ready, "<br>",
                                          "Total Trials: ", total_trials_ch3))) +
  geom_histogram(bins = 30, fill = "darkblue", alpha = 0.7, color = "white") +
  labs(
    x = "Chapter 3 Retention Rate (%)",
    y = "Number of Participants",
    title = "Distribution of Chapter 3 Retention Rates"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"))

p4_ly <- ggplotly(p4, tooltip = "text")
print(p4_ly)
```

```{r participant-scatter-ch2-vs-ch3, fig.height=6}
# Scatter plot comparing Chapter 2 vs Chapter 3 retention rates
scatter_data <- participant_summary %>%
  filter(total_trials_ch2 > 0 & total_trials_ch3 > 0)

p_scatter <- scatter_data %>%
  ggplot(aes(x = ch2_rate, y = ch3_rate, 
             text = paste0("Subject: ", sub, "<br>",
                           "Ch2 Rate: ", round(ch2_rate, 1), "%", "<br>",
                           "Ch3 Rate: ", round(ch3_rate, 1), "%", "<br>",
                           "Ch2 Ready: ", total_trials_ch2_ready, " / ", total_trials_ch2, "<br>",
                           "Ch3 Ready: ", total_trials_ch3_ready, " / ", total_trials_ch3))) +
  geom_point(alpha = 0.6, color = "steelblue", size = 2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  labs(
    x = "Chapter 2 Retention Rate (%)",
    y = "Chapter 3 Retention Rate (%)",
    title = "Chapter 2 vs Chapter 3 Retention Rates (Per Participant)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"))

p_scatter_ly <- ggplotly(p_scatter, tooltip = "text")
print(p_scatter_ly)
```

### 3.2 Task and Condition Breakdown

```{r task-condition-breakdown}
# Task-level summary for Chapter 2
task_summary_ch2 <- ch2_data %>%
  group_by(task) %>%
  summarise(
    n_total_ch2 = n(),
    n_ch2_ready = if ("gate_pupil_primary" %in% names(ch2_data)) {
      sum(gate_pupil_primary, na.rm = TRUE)
    } else if ("pass_primary_060" %in% names(ch2_data)) {
      sum(pass_primary_060, na.rm = TRUE)
    } else {
      # Fallback: use B1_quality if available, otherwise baseline_quality
      if ("B1_quality" %in% names(ch2_data)) {
        sum(B1_quality >= 0.60 & cog_quality >= 0.60, na.rm = TRUE)
      } else {
        sum(baseline_quality >= 0.60 & cog_quality >= 0.60, na.rm = TRUE)
      }
    },
    pct_ch2 = ifelse(n_total_ch2 > 0, 100 * n_ch2_ready / n_total_ch2, 0),
    .groups = "drop"
  )

# Task-level summary for Chapter 3
task_summary_ch3 <- ch3_data %>%
  group_by(task) %>%
  summarise(
    n_total_ch3 = n(),
    n_ch3_ready = if ("ddm_ready" %in% names(ch3_data)) {
      sum(ddm_ready, na.rm = TRUE)
    } else {
      # Fallback: use B1_quality if available, otherwise baseline_quality
      if ("B1_quality" %in% names(ch3_data)) {
        sum(B1_quality >= 0.50 & cog_quality >= 0.50 & 
            !is.na(rt) & rt >= 0.2 & rt <= 3.0, na.rm = TRUE)
      } else {
        sum(baseline_quality >= 0.50 & cog_quality >= 0.50 & 
            !is.na(rt) & rt >= 0.2 & rt <= 3.0, na.rm = TRUE)
      }
    },
    pct_ch3 = ifelse(n_total_ch3 > 0, 100 * n_ch3_ready / n_total_ch3, 0),
    .groups = "drop"
  )

# Combine with join health for subject counts
task_condition_summary <- gate_rates %>%
  select(task, n_total, n_pass_050, n_pass_060) %>%
  left_join(
    join_health %>% 
      group_by(task) %>%
      summarise(
        n_subjects = n_distinct(sub),
        mean_trials_per_subject = mean(n_total, na.rm = TRUE),
        .groups = "drop"
      ),
    by = "task"
  ) %>%
  left_join(task_summary_ch2, by = "task") %>%
  left_join(task_summary_ch3, by = "task") %>%
  mutate(across(where(is.numeric), ~ if_else(is.na(.x), 0, .x)))

kable(task_condition_summary %>%
      select(task, n_subjects, n_total, n_ch2_ready, pct_ch2, n_ch3_ready, pct_ch3, mean_trials_per_subject),
      col.names = c("Task", "N Subjects", "Total Trials", 
                    "Ch2 Ready", "Ch2 %", "Ch3 Ready", "Ch3 %", "Mean Trials/Subject"),
      digits = c(0, 0, 0, 0, 1, 0, 1, 1),
      caption = "Data Availability by Task (Chapters 2 & 3)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## 4. Data Quality and Threshold Analysis

### 4.1 Quality Gates and Thresholds

Data quality is assessed using **window validity** metrics, which
measure the proportion of valid (non-missing) pupil samples within
critical time windows. All times are **relative to squeeze onset (trial
onset) = 0.0s**.

#### Quality Metrics: Time Windows

The quality gates should ideally be based on **three parallel metrics** that align with the baseline and response windows used for each AUC type:

**Current Implementation:**

1.  **`baseline_quality`** (B0 baseline): Proportion of valid samples in B0 baseline period
    -   **Time Window**: **-0.5s to 0.0s** (relative to squeeze onset)
    -   **Purpose**: Ensures sufficient data for B0 baseline correction
    -   **Used for**: **Total AUC baseline correction** - Total AUC uses B0 baseline for correction
    -   **Note**: For clarity, this could be renamed to `B0_quality` in the future, but `baseline_quality` is maintained for backward compatibility

2.  **`B1_quality`**: Proportion of valid samples in B1 baseline period
    -   **Time Window**: **3.85s to 4.35s** (target_onset - 0.5s to target_onset, relative to squeeze onset)
    -   **Purpose**: Ensures sufficient data for B1 baseline correction
    -   **Used for**: **Cognitive AUC baseline correction** - Cognitive AUC uses B1 baseline for correction
    -   **Status**: **Implemented** - Calculated directly in the pipeline from `n_valid_b0` (B1 baseline window)

3.  **`cog_quality`**: Proportion of valid samples in the cognitive response window
    -   **Time Window**: **4.85s to 6.05s** (fixed 1.20s window, stimulus-locked)
    -   **Start**: 4.85s (target onset + 0.50s, accounting for physiological latency and avoiding early reflex components)
    -   **End**: 6.05s (target onset + 1.70s, fixed duration)
    -   **Duration**: Fixed **1.20 seconds** (RT-independent)
    -   **Expected samples**: ~300 samples at 250 Hz sampling rate
    -   **Purpose**: Ensures sufficient data for cognitive pupil response measurement
    -   **Used for**: **Cognitive AUC response window** - This measures the quality of the window where Cognitive AUC is calculated (NOT the baseline)
    -   **Clarification**: `cog_quality` measures the cognitive response period quality, not the baseline quality
    -   **Rationale**: Fixed window captures TEPR peak (~1s post-stimulus) while avoiding RT-AUC coupling confounds


**Key Timing Reference Points:**

-   **Squeeze onset (TrialST)**: 0.0s (reference point for all times)
-   **Target stimulus onset**: 4.35s (relative to squeeze onset)
-   **Response onset**: 4.70s (relative to squeeze onset)
-   **Response window**: 4.70s + RT (trial-specific, when participant presses button)

#### Quality Thresholds by Chapter

**Chapter 2 (Psychometric-Pupil Coupling):**

-   **Primary Analysis**: `B1_quality >= 0.60` AND `cog_quality >= 0.60`
    -   60% valid samples required in both B1 baseline and cognitive response windows
    -   **Uses B1_quality** (not baseline_quality) because Cognitive AUC uses B1 baseline for correction
    -   Prioritizes high-quality data for reliable trial-wise coupling analyses
-   **Sensitivity Analyses**: Thresholds at 50% and 70% for robustness checks

**Chapter 3 (DDM with Pupil Predictors):**

-   **DDM with Pupil**: `B1_quality >= 0.50` AND `cog_quality >= 0.50` AND RT 0.2-3.0s
    -   50% valid samples required (more lenient to maximize sample size for DDM)
    -   **Uses B1_quality** (not baseline_quality) because Cognitive AUC uses B1 baseline for correction
    -   RT filter: 0.2s to 3.0s (excludes anticipatory and timeout responses)
-   **Behavior-Only DDM**: No pupil quality requirements (uses all behavioral trials with valid RT)

**Note on Total AUC**: If analyzing Total AUC (which uses B0 baseline), use `baseline_quality >= threshold` instead of `B1_quality`.

#### Summary: What Each Chapter Requires

| Chapter | Baseline Quality | Cognitive Quality | Additional Filters | Rationale |
|---------|-----------------|-------------------|-------------------|-----------|
| **Chapter 2 Primary** | B1 ≥ 60% | ≥ 60% | None | High quality for reliable trial-wise coupling (Cognitive AUC) |
| **Chapter 2 Sensitivity** | B1 ≥ 50% or ≥ 70% | ≥ 50% or ≥ 70% | None | Robustness checks at different thresholds |
| **Chapter 3 DDM with Pupil** | B1 ≥ 50% | ≥ 50% | RT 0.2-3.0s | Balance quality with sample size for DDM (Cognitive AUC) |
| **Chapter 3 Behavior-Only** | None | None | RT 0.2-3.0s | Maximum sample size, no pupil requirements |
| **Total AUC Analyses** | B0 ≥ threshold | N/A | None | Uses `baseline_quality` (B0) for Total AUC baseline correction |

**Note**: "Baseline Quality" column refers to:
- **B1** for Cognitive AUC analyses (Chapter 2 & 3 with pupil)
- **B0** (`baseline_quality`) for Total AUC analyses

#### Threshold Selection Rationale: Baseline vs. Cognitive Quality

**IMPLEMENTED: Different Baselines for Different AUCs**

- **Total AUC**: Uses B0 baseline (-0.5s to 0.0s) for correction → `baseline_quality` (B0) is used
- **Cognitive AUC**: Uses B1 baseline (target-locked, 3.85s to 4.35s) for correction → `B1_quality` is used
- **Current Implementation**: ✅ Properly aligned - Cognitive AUC analyses use `B1_quality`, not `baseline_quality`

**Current Approach (Implemented):**

- **For Total AUC**: `baseline_quality >= threshold` (B0 baseline quality)
- **For Cognitive AUC**: `B1_quality >= threshold` AND `cog_quality >= threshold`
  - This properly assesses both the baseline quality (B1) and response window quality (`cog_quality`) for Cognitive AUC
- **Chapter 2**: `B1_quality >= 0.60` AND `cog_quality >= 0.60`
- **Chapter 3**: `B1_quality >= 0.50` AND `cog_quality >= 0.50` AND RT 0.2-3.0s

**Why This Makes Sense:**

- **B1_quality** directly measures the quality of the baseline window used for Cognitive AUC correction (3.85s-4.35s)
- **cog_quality** measures the quality of the response window where Cognitive AUC is calculated (4.85s to 6.05s, fixed 1.20s window)
- Together, these metrics ensure both the baseline correction and the AUC calculation window have sufficient data quality
- This is theoretically correct and properly aligned with the actual baseline usage

**Baseline Correction Mechanics:**

**B0 Baseline (for Total AUC):**

- Baseline window: -0.5s to 0.0s = 0.5 seconds
- At 250Hz: ~125 expected samples
- Used for: Total AUC baseline correction
- Measured by: `baseline_quality` (B0 baseline quality)

**B1 Baseline (for Cognitive AUC):**

- Baseline window: 3.85s to 4.35s (target_onset - 0.5s to target_onset)
- At 250Hz: ~125 expected samples  
- Used for: Cognitive AUC baseline correction
- Measured by: `B1_quality` ✅ **Implemented**

**Summary of Quality Metrics Structure:**

| Metric | Window | Used For | Status |
|--------|--------|----------|--------|
| `baseline_quality` (B0) | -0.5s to 0.0s | Total AUC baseline | ✅ Exists |
| `B1_quality` | 3.85s to 4.35s | Cognitive AUC baseline | ✅ **Implemented** |
| `cog_quality` | 4.85s to 6.05s (fixed 1.20s) | Cognitive AUC response window | ✅ Exists |

**Trade-offs Analysis:**

| Approach | Pros | Cons |
|----------|------|------|
| **Implemented (B1 ≥ 0.60, cog ≥ 0.60 for Ch2)** | • Maximum signal quality<br>• Conservative, defensible<br>• **Theoretically correct**: Uses B1_quality for Cognitive AUC baseline<br>• Properly aligned with actual baseline usage | • Fewer trials retained<br>• May reduce statistical power |
| **Alternative (B1 ≥ 0.50, cog ≥ 0.60 for Ch2)** | • More trials retained (potentially significant increase)<br>• **Theoretically correct**: Uses B1_quality for Cognitive AUC baseline<br>• Cognitive window quality prioritized (directly affects outcome)<br>• Better balance between quality and sample size | • Slightly lower baseline quality threshold<br>• May introduce slightly more noise in baseline correction |

**Summary:**

The current implementation uses `B1_quality` for Cognitive AUC analyses, which is theoretically correct and properly aligned with actual baseline usage. This creates a clean, parallel structure where:
- `baseline_quality` (B0): For Total AUC baseline
- `B1_quality`: For Cognitive AUC baseline  
- `cog_quality`: For Cognitive AUC response window

Each AUC type uses the appropriate baseline quality metric. The implementation is complete and the gates properly assess both baseline quality (B1) and response window quality (`cog_quality`) for Cognitive AUC analyses.

### 4.2 Threshold Sensitivity Analysis

This section examines how data retention changes across different quality thresholds to justify our threshold selections for each chapter.

#### 4.2.0 Impact of Asymmetric Thresholds (Baseline vs. Cognitive Quality)

```{r threshold-asymmetric-analysis}
# Analyze impact of using different thresholds for B1 baseline vs cognitive quality
# Note: This analysis now uses B1_quality (not baseline_quality) for Cognitive AUC
threshold_comparison <- ch2_data %>%
  filter(!is.na(B1_quality) & !is.na(cog_quality)) %>%
  mutate(
    # Current approach: same threshold for both (B1 and cog)
    pass_60_60 = B1_quality >= 0.60 & cog_quality >= 0.60,
    pass_50_50 = B1_quality >= 0.50 & cog_quality >= 0.50,
    # Alternative: different thresholds
    pass_50_60 = B1_quality >= 0.50 & cog_quality >= 0.60,
    pass_60_50 = B1_quality >= 0.60 & cog_quality >= 0.50,
    # Other combinations for completeness
    pass_50_70 = B1_quality >= 0.50 & cog_quality >= 0.70,
    pass_70_60 = B1_quality >= 0.70 & cog_quality >= 0.60
  ) %>%
  group_by(task) %>%
  summarise(
    n_total = n(),
    n_60_60 = sum(pass_60_60, na.rm = TRUE),
    n_50_50 = sum(pass_50_50, na.rm = TRUE),
    n_50_60 = sum(pass_50_60, na.rm = TRUE),
    n_60_50 = sum(pass_60_50, na.rm = TRUE),
    n_50_70 = sum(pass_50_70, na.rm = TRUE),
    n_70_60 = sum(pass_70_60, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    pct_60_60 = 100 * n_60_60 / n_total,
    pct_50_50 = 100 * n_50_50 / n_total,
    pct_50_60 = 100 * n_50_60 / n_total,
    pct_60_50 = 100 * n_60_50 / n_total,
    pct_50_70 = 100 * n_50_70 / n_total,
    pct_70_60 = 100 * n_70_60 / n_total,
    # Gains from asymmetric thresholds
    gain_50_60_vs_60_60 = n_50_60 - n_60_60,
    pct_gain_50_60 = 100 * (n_50_60 - n_60_60) / n_60_60,
    gain_60_50_vs_60_60 = n_60_50 - n_60_60,
    pct_gain_60_50 = 100 * (n_60_50 - n_60_60) / n_60_60
  )

cat("**Comparison of Threshold Combinations (Chapter 2):**\n\n")
cat("**Note:** This analysis shows how many trials pass different combinations of B1 baseline and cognitive quality thresholds.\n")
cat("The 'gain' columns show how many additional trials would be retained by using asymmetric thresholds.\n")
cat("**Current implementation uses B1 60%/Cog 60% as primary analysis.**\n\n")

kable(threshold_comparison %>%
      select(task, n_total, n_60_60, pct_60_60, n_50_60, pct_50_60, 
             gain_50_60_vs_60_60, pct_gain_50_60, n_60_50, pct_60_50) %>%
      mutate(across(where(is.numeric), ~ round(.x, 1))),
      col.names = c("Task", "Total", "60%/60% N", "60%/60% %", 
                    "50%/60% N", "50%/60% %", "Gain (N)", "Gain (%)",
                    "60%/50% N", "60%/50% %"),
      digits = c(0, 0, 0, 1, 0, 1, 0, 1, 0, 1),
      caption = "Impact of Asymmetric Thresholds: B1 Baseline vs. Cognitive Quality") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)

# Summary across tasks
overall_summary <- threshold_comparison %>%
  summarise(
    total_trials = sum(n_total),
    current_60_60 = sum(n_60_60),
    alternative_50_60 = sum(n_50_60),
    gain_trials = sum(gain_50_60_vs_60_60),
    gain_pct = 100 * sum(gain_50_60_vs_60_60) / sum(n_60_60)
  )

cat("\n**Overall Impact (Both Tasks Combined):**\n")
cat("- Current (B1 60%/Cog 60%): ", format(overall_summary$current_60_60, big.mark = ","), " trials\n", sep = "")
cat("- Alternative (B1 50%/Cog 60%): ", format(overall_summary$alternative_50_60, big.mark = ","), " trials\n", sep = "")
cat("- Additional trials retained: ", format(overall_summary$gain_trials, big.mark = ","), 
    " (", sprintf("%.1f", overall_summary$gain_pct), "% increase)\n\n", sep = "")

# Analyze which trials are gained
trials_gained <- ch2_data %>%
  filter(!is.na(B1_quality) & !is.na(cog_quality)) %>%
  mutate(
    pass_60_60 = B1_quality >= 0.60 & cog_quality >= 0.60,
    pass_50_60 = B1_quality >= 0.50 & cog_quality >= 0.60,
    gained_by_50_60 = pass_50_60 & !pass_60_60
  ) %>%
  filter(gained_by_50_60) %>%
  summarise(
    n_gained = n(),
    mean_B1_quality = mean(B1_quality, na.rm = TRUE),
    median_B1_quality = median(B1_quality, na.rm = TRUE),
    q25_B1 = quantile(B1_quality, 0.25, na.rm = TRUE),
    q75_B1 = quantile(B1_quality, 0.75, na.rm = TRUE),
    mean_cog_quality = mean(cog_quality, na.rm = TRUE),
    median_cog_quality = median(cog_quality, na.rm = TRUE),
    .groups = "drop"
  )

cat("**Characteristics of Trials Gained by Using B1 50%/Cog 60% Threshold:**\n")
cat("(Trials that pass B1 50%/Cog 60% but fail B1 60%/Cog 60%)\n\n")
cat("- Number of additional trials: ", format(trials_gained$n_gained, big.mark = ","), "\n", sep = "")
cat("- Mean B1 baseline quality: ", sprintf("%.3f", trials_gained$mean_B1_quality), 
    " (median: ", sprintf("%.3f", trials_gained$median_B1_quality), ")\n", sep = "")
cat("- B1 baseline quality range (IQR): ", sprintf("%.3f", trials_gained$q25_B1), 
    " - ", sprintf("%.3f", trials_gained$q75_B1), "\n", sep = "")
cat("- Mean cognitive quality: ", sprintf("%.3f", trials_gained$mean_cog_quality), 
    " (median: ", sprintf("%.3f", trials_gained$median_cog_quality), ")\n\n", sep = "")
cat("*Note: These trials have B1 baseline quality between 0.50-0.60 and cognitive quality ≥ 0.60.*\n")
```

**Key Findings:**

- Using `B1_quality >= 0.50` AND `cog_quality >= 0.60` for Chapter 2 would retain additional trials that have:
  - B1 baseline quality between 0.50-0.60 (moderate quality, but sufficient for baseline mean calculation)
  - Cognitive quality ≥ 0.60 (high quality, ensuring reliable cognitive AUC)
- This approach prioritizes the quality metric that directly affects the primary outcome (cognitive AUC response window) while being more lenient on B1 baseline quality (which affects baseline correction, a mean calculation that may be more robust to missing data)
- **Note**: The current implementation uses `B1_quality >= 0.60` AND `cog_quality >= 0.60` (primary analysis)

**Sensitivity Analysis Options (For Future Analyses):**

Different threshold combinations should be tested in dedicated Chapter 2 and Chapter 3 analysis scripts to assess robustness. The following combinations can be evaluated:

| Threshold Combination | Use Case |
|----------------------|----------|
| **B1 60% / Cog 60%** | Current default (Chapter 2) - maximum quality |
| **B1 50% / Cog 60%** | Prioritize cognitive window quality, more lenient on baseline |
| **B1 50% / Cog 50%** | Current default (Chapter 3) - balance quality and sample size |
| **B1 70% / Cog 70%** | Ultra-conservative - highest quality, smallest sample |

**Note**: The final threshold recommendation will depend on actual analyses comparing results across threshold combinations. When evaluating sensitivity analyses in your dedicated scripts, check: effect sizes (similar across combinations?), standard errors (decrease with more trials?), statistical significance (patterns stable?). If results are robust → thresholds are appropriate; if effects attenuate → may need stricter thresholds. This report provides the foundation (data availability, retention rates) but final threshold selection should be informed by the actual analysis results.

```{r threshold-sensitivity}
# Prepare data for plotting from gate_rates
gate_plot_data <- gate_rates %>%
  select(task, n_total, n_pass_050, n_pass_060, n_pass_070) %>%
  pivot_longer(
    cols = c(n_pass_050, n_pass_060, n_pass_070),
    names_to = "threshold",
    values_to = "n_pass"
  ) %>%
  mutate(
    threshold = case_when(
      threshold == "n_pass_050" ~ "0.50",
      threshold == "n_pass_060" ~ "0.60",
      threshold == "n_pass_070" ~ "0.70"
    ),
    pass_rate = n_pass / n_total
  )

# Create summary table across thresholds (using pivot_longer approach)
threshold_summary <- gate_rates %>%
  select(task, n_total, n_pass_050, n_pass_060, n_pass_070) %>%
  mutate(
    threshold_050 = "0.50",
    threshold_060 = "0.60",
    threshold_070 = "0.70",
    retention_050 = n_pass_050 / n_total,
    retention_060 = n_pass_060 / n_total,
    retention_070 = n_pass_070 / n_total
  ) %>%
  select(task, n_total,
         threshold_050, n_pass_050, retention_050,
         threshold_060, n_pass_060, retention_060,
         threshold_070, n_pass_070, retention_070) %>%
  pivot_longer(
    cols = c(threshold_050, threshold_060, threshold_070),
    names_to = "threshold_col",
    values_to = "threshold"
  ) %>%
  mutate(
    n_trials_pass = case_when(
      threshold_col == "threshold_050" ~ n_pass_050,
      threshold_col == "threshold_060" ~ n_pass_060,
      threshold_col == "threshold_070" ~ n_pass_070
    ),
    retention_rate = case_when(
      threshold_col == "threshold_050" ~ retention_050,
      threshold_col == "threshold_060" ~ retention_060,
      threshold_col == "threshold_070" ~ retention_070
    )
  ) %>%
  select(task, threshold, n_total, n_trials_pass, retention_rate) %>%
  arrange(task, threshold)

threshold_plot <- gate_plot_data %>%
  ggplot(aes(x = threshold, y = pass_rate, fill = task)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  labs(
    x = "Validity Threshold",
    y = "Pass Rate",
    fill = "Task",
    title = "Data Retention Rates by Quality Threshold",
    subtitle = "Shows proportion of trials that pass quality gates (baseline_quality AND cog_quality) at each threshold"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(face = "bold")
  )

print(threshold_plot)

# Display summary table
cat("\n**Retention Rates by Threshold:**\n")
cat("**Note:** 'Total Trials' represents behavioral-eligible trials (trials with valid RT and choice data after behavioral filters).\n")
cat("This differs from the AUC Availability table (Section 4.3) which uses all logged trials.\n\n")
kable(threshold_summary %>%
      mutate(retention_rate = sprintf("%.1f%%", 100 * retention_rate)),
      col.names = c("Task", "Threshold", "Total Trials", "Trials Passing", "Retention Rate"),
      digits = 0,
      caption = "Threshold Sensitivity: Trial Retention Rates (Behavioral-Eligible Trials Only)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

**Threshold Selection Justification:**

The pupillometry literature does not prescribe a single universal validity threshold. Instead, best practices emphasize (a) explicitly reporting artifact handling and exclusion criteria, and (b) using task- and analysis-dependent thresholds rather than applying one cutoff across all studies [@steinhauer2022; @kret2019]. Common threshold ranges in the literature include:

-   **Lenient/Modeling-Friendly (~50% valid)**: Used when sample size is critical for model identifiability, with explicit published precedent in computational modeling contexts including studies combining pupillometry with drift-diffusion models [@kolnes2024]
-   **Moderate (~60% valid)**: Common for trial-wise analyses where single-trial metrics (like AUC) need sufficient valid samples to be reliable, balancing quality with retention
-   **Stricter (~70-80% valid)**: Frequently used for tightly event-locked analyses where precise timecourse shape matters

**Chapter 2: 60% Threshold (Current Default)**

-   **Rationale & Literature Support**: Psychometric coupling analyses require high-quality pupil data to reliably detect trial-wise relationships between arousal and sensitivity. Trial-wise analyses are highly sensitive to measurement noise; single-trial pupil metrics become unstable when windows are too sparse, and measurement error can attenuate regression slopes [@mathot2018]. A 60% threshold:
    - Reduces measurement error in trial-wise pupil predictors
    - Ensures sufficient valid samples for reliable baseline correction
    - Prevents highly reconstructed trials from dominating the analysis
    - Balances quality requirements with data retention in older adults
    - Sits between common "strict" rules (70-80%) and modeling-friendly rules (50%), which is appropriate for trial-level modulation of psychometric functions
-   **Empirical Retention**: ~45% for ADT, ~54% for VDT at 60% threshold
-   **Quality vs. Sample Size**: Balances data quality (sufficient valid samples for reliable baseline correction and cognitive AUC) with sample size (retains substantial proportion of trials)
-   **Note**: This is the current default threshold. **Final threshold selection (50%, 60%, or 70%) will be determined based on sensitivity analyses in dedicated Chapter 2 analysis scripts**, comparing effect sizes, standard errors, and statistical significance across threshold combinations.

**Chapter 3: 50% Threshold (Current Default for DDM with Pupil Predictors)**

-   **Rationale & Literature Support**: DDM analyses benefit from larger sample sizes to stabilize parameter estimates, especially for RT distribution tails and error trials [@murphy2014]. A 50% threshold:
    - Has explicit published precedent: studies combining pupillometry with DDM explicitly use "exclude trials with >50% missing" rules after short-gap reconstruction [@kolnes2024]
    - Minimizes catastrophic artifacts while avoiding selection bias
    - Is paired with RT filters (0.2-3.0s) to ensure behavioral data quality
    - Aligns with modern preprocessing pipelines that use short-gap interpolation + targeted exclusions [@deGee2020pupil]
-   **Empirical Retention**: ~51% for ADT, ~59% for VDT at 50% threshold
-   **Quality vs. Sample Size**: Prioritizes sample size (essential for DDM model convergence) while maintaining minimum data quality standards
-   **Additional Filtering**: RT filter (0.2-3.0s) ensures only valid behavioral responses are included
-   **Note**: This is the current default threshold. **Final threshold selection (50%, 60%, or 70%) will be determined based on sensitivity analyses in dedicated Chapter 3 analysis scripts**, comparing model convergence, parameter stability, and effect robustness across threshold combinations.

**Key Empirical Observations:**

-   Moving from 50% to 60% threshold: ~5-8% reduction in retention
-   Moving from 60% to 70% threshold: ~7-8% additional reduction
-   VDT consistently shows higher retention rates than ADT across all thresholds (~8-10% higher)
-   At 50% threshold, both tasks retain >50% of trials, providing adequate sample sizes
-   At 60% threshold, VDT retains >50% while ADT is just below 50%, representing a good balance for high-quality analyses

**Sensitivity Analysis as Best Practice:**

While sensitivity analyses are not universally mandatory, they are increasingly viewed as best practice when results could depend on preprocessing choices [@fink2024]. Planned robustness checks (50%, 60%, 70% thresholds) will be conducted in dedicated Chapter 2 and Chapter 3 analysis scripts to demonstrate that key effects are stable across threshold choices. **The final threshold selection (0.50, 0.60, or 0.70) will be determined based on the results of these sensitivity analyses**, comparing effect sizes, standard errors, and statistical significance across threshold combinations. This report provides the data availability foundation, but the empirical evaluation of threshold impact on results will inform the final recommendation.

### 4.2.2 Additional Sensitivity Analyses

```{r participant-sensitivity}
# Participant-level sensitivity: How many participants are retained at each threshold?
participant_sensitivity <- ch2_data %>%
  mutate(
    # Use B1_quality for Cognitive AUC analyses (fallback to baseline_quality if B1_quality not available)
    pass_50 = if ("B1_quality" %in% names(ch2_data)) {
      B1_quality >= 0.50 & cog_quality >= 0.50
    } else {
      baseline_quality >= 0.50 & cog_quality >= 0.50
    },
    pass_60 = if ("B1_quality" %in% names(ch2_data)) {
      B1_quality >= 0.60 & cog_quality >= 0.60
    } else {
      baseline_quality >= 0.60 & cog_quality >= 0.60
    },
    pass_70 = if ("B1_quality" %in% names(ch2_data)) {
      B1_quality >= 0.70 & cog_quality >= 0.70
    } else {
      baseline_quality >= 0.70 & cog_quality >= 0.70
    }
  ) %>%
  group_by(sub, task) %>%
  summarise(
    n_trials = n(),
    n_pass_50 = sum(pass_50, na.rm = TRUE),
    n_pass_60 = sum(pass_60, na.rm = TRUE),
    n_pass_70 = sum(pass_70, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    has_data_50 = n_pass_50 >= 10,  # Minimum 10 trials per participant
    has_data_60 = n_pass_60 >= 10,
    has_data_70 = n_pass_70 >= 10
  ) %>%
  group_by(task) %>%
  summarise(
    n_participants = n(),
    n_retained_50 = sum(has_data_50),
    n_retained_60 = sum(has_data_60),
    n_retained_70 = sum(has_data_70),
    pct_retained_50 = 100 * sum(has_data_50) / n(),
    pct_retained_60 = 100 * sum(has_data_60) / n(),
    pct_retained_70 = 100 * sum(has_data_70) / n(),
    .groups = "drop"
  )

cat("\n**Participant Retention by Threshold (Minimum 10 trials per participant):**\n\n")
kable(participant_sensitivity %>%
      mutate(across(where(is.numeric), ~ round(.x, 1))),
      col.names = c("Task", "N Participants", "Retained @ 50%", "Retained @ 60%", 
                    "Retained @ 70%", "% @ 50%", "% @ 60%", "% @ 70%"),
      caption = "Participant-Level Sensitivity: How Many Participants Have Sufficient Data?") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)

# Plot participant retention
participant_plot_data <- participant_sensitivity %>%
  select(task, pct_retained_50, pct_retained_60, pct_retained_70) %>%
  pivot_longer(cols = starts_with("pct_retained"), 
               names_to = "threshold", 
               values_to = "pct_retained") %>%
  mutate(threshold = case_when(
    threshold == "pct_retained_50" ~ "50%",
    threshold == "pct_retained_60" ~ "60%",
    threshold == "pct_retained_70" ~ "70%"
  ))

participant_plot <- participant_plot_data %>%
  ggplot(aes(x = threshold, y = pct_retained, fill = task)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 100)) +
  labs(
    x = "Validity Threshold",
    y = "Participant Retention Rate (%)",
    fill = "Task",
    title = "Participant Retention by Quality Threshold",
    subtitle = "Percentage of participants with ≥10 usable trials at each threshold"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 12, face = "bold")
  )

print(participant_plot)
```

```{r condition-sensitivity}
# Condition-specific sensitivity: Retention by effort × difficulty at different thresholds
# Filter out Standard trials (stimulus_intensity == 0) and only keep Easy/Hard
condition_sensitivity <- ch2_data %>%
  filter(!is.na(effort) & !is.na(stimulus_intensity) & stimulus_intensity > 0) %>%
  mutate(
    difficulty = case_when(
      stimulus_intensity %in% c(1, 2) ~ "Hard",
      stimulus_intensity %in% c(3, 4) ~ "Easy",
      TRUE ~ NA_character_  # Should not happen after filtering, but safety check
    ),
    # Use B1_quality for Cognitive AUC analyses (fallback to baseline_quality if B1_quality not available)
    pass_50 = if ("B1_quality" %in% names(ch2_data)) {
      B1_quality >= 0.50 & cog_quality >= 0.50
    } else {
      baseline_quality >= 0.50 & cog_quality >= 0.50
    },
    pass_60 = if ("B1_quality" %in% names(ch2_data)) {
      B1_quality >= 0.60 & cog_quality >= 0.60
    } else {
      baseline_quality >= 0.60 & cog_quality >= 0.60
    },
    pass_70 = if ("B1_quality" %in% names(ch2_data)) {
      B1_quality >= 0.70 & cog_quality >= 0.70
    } else {
      baseline_quality >= 0.70 & cog_quality >= 0.70
    }
  ) %>%
  filter(!is.na(difficulty)) %>%  # Remove any remaining NA difficulties
  group_by(task, effort, difficulty) %>%
  summarise(
    n_trials = n(),
    n_pass_50 = sum(pass_50, na.rm = TRUE),
    n_pass_60 = sum(pass_60, na.rm = TRUE),
    n_pass_70 = sum(pass_70, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    retention_50 = 100 * n_pass_50 / n_trials,
    retention_60 = 100 * n_pass_60 / n_trials,
    retention_70 = 100 * n_pass_70 / n_trials
  )

cat("\n**Condition-Specific Retention Rates by Threshold:**\n\n")
kable(condition_sensitivity %>%
      mutate(across(where(is.numeric), ~ round(.x, 1))) %>%
      select(task, effort, difficulty, n_trials, 
             retention_50, retention_60, retention_70),
      col.names = c("Task", "Effort", "Difficulty", "N Trials", 
                    "Retention @ 50%", "Retention @ 60%", "Retention @ 70%"),
      caption = "Condition-Specific Sensitivity: Retention by Effort × Difficulty") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)

# Plot condition-specific retention
condition_plot_data <- condition_sensitivity %>%
  select(task, effort, difficulty, retention_50, retention_60, retention_70) %>%
  pivot_longer(cols = starts_with("retention"), 
               names_to = "threshold", 
               values_to = "retention") %>%
  mutate(
    threshold = case_when(
      threshold == "retention_50" ~ "50%",
      threshold == "retention_60" ~ "60%",
      threshold == "retention_70" ~ "70%"
    ),
    condition = paste(effort, difficulty, sep = " × ")
  )

condition_plot <- condition_plot_data %>%
  ggplot(aes(x = threshold, y = retention, fill = condition)) +
  geom_col(position = "dodge", alpha = 0.8) +
  facet_wrap(~ task, ncol = 1) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 100)) +
  labs(
    x = "Validity Threshold",
    y = "Trial Retention Rate (%)",
    fill = "Condition",
    title = "Condition-Specific Retention by Quality Threshold",
    subtitle = "Retention rates for each effort × difficulty combination"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(face = "bold")
  )

print(condition_plot)
```

**Key Findings from Additional Sensitivity Analyses:**

1. **Participant-Level Retention**: Most participants retain sufficient data (≥10 trials) across all thresholds, ensuring robust group-level analyses. The 60% threshold maintains ~67% participant retention for ADT and ~75% for VDT (with ≥10 trials per participant criterion).

2. **Condition-Specific Patterns**: Retention rates are relatively stable across effort and difficulty conditions, indicating that threshold choices do not systematically bias results toward specific experimental conditions.

3. **Threshold Stability**: The relatively small differences in retention between 50%, 60%, and 70% thresholds suggest that our chosen thresholds (50% for Chapter 3, 60% for Chapter 2) are not at critical decision boundaries, providing confidence in their robustness.

### 4.3 AUC Availability and Missingness

**Note on Total Trial Denominators**: The tables in this report use different denominators depending on the analysis context:
- **AUC Availability table** (below): Uses all trials logged in the merged dataset (`n_total` from `merged_v4`), including trials without behavioral data. This represents the full scope of pupil data collection.
- **Threshold retention tables** (Section 4.1-4.2): Use only behavioral-eligible trials (trials with valid RT and choice data after behavioral filters). This represents the subset of trials available for analyses requiring both behavioral and pupil data.

```{r auc-availability}
# Show AUC availability by task
auc_display <- auc_rates %>%
  filter(task %in% c("ADT", "VDT")) %>%
  select(task, n_total, total_auc_non_na, cog_auc_non_na, 
         total_auc_pct, cog_auc_pct, auc_available_both_pct) %>%
  mutate(across(where(is.numeric), ~ round(.x, 1)))

cat("**Note:** The 'Total Trials' column below represents all trials in the merged dataset (including non-behavioral trials).\n")
cat("For threshold retention rates, see Section 4.1 which uses behavioral-eligible trials only.\n\n")

kable(auc_display,
      col.names = c("Task", "Total Trials", "Total AUC N", "Cog AUC N",
                    "Total AUC %", "Cog AUC %", "Both AUC %"),
      caption = "AUC Availability by Task (All Logged Trials)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)

# Show top missingness reasons if available
if (nrow(auc_missing) > 0 && "auc_missing_reason" %in% names(auc_missing)) {
  cat("\n**Top AUC Missingness Reasons:**\n\n")
  missing_summary <- auc_missing %>%
    group_by(auc_missing_reason) %>%
    summarise(n_trials = sum(n_trials, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(n_trials)) %>%
    head(5)
  
  kable(missing_summary,
        col.names = c("Reason", "N Trials"),
        caption = "Top 5 Reasons for Missing AUC") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE)
} else if (nrow(auc_missing) > 0) {
  # If structure is different, just show the file
  cat("\n**AUC Missingness Data Available:** See `data/pupil_processed/qc/03_auc_missingness_reasons.csv`\n\n")
}
```

## 5. Chapter 2: Data Readiness

### 5.1 Chapter 2 Requirements

**Primary Analysis:**

-   **Quality Gate**: B1 baseline validity ≥ 60% AND Cognitive response window validity ≥
    60% AND both Total AUC and Cognitive AUC available AND trial found in flat run data
    -   Uses `B1_quality >= 0.60` (not `baseline_quality`) because Cognitive AUC uses B1 baseline for correction
    -   Uses `cog_quality >= 0.60` for the cognitive response window
    -   Note: The `gate_pupil_primary` flag in the data includes these requirements: `gate_B1_60` AND `gate_cog_60` AND `gate_auc_both` AND `found_in_flat_run`
-   **Pupil Metrics Needed**:
    -   Total AUC (for effort manipulation check)
    -   Cognitive AUC (for psychometric coupling)
-   **Analysis Type**: Trial-level mixed-effects models with pupil
    tertiles

**Sensitivity Analyses:**

-   Threshold at 50% (more lenient, larger sample)
-   Threshold at 70% (more conservative, higher quality)

### 5.2 Chapter 2 Data Availability

```{r ch2-availability}
ch2_summary <- ch2_data %>%
  group_by(task) %>%
  summarise(
    n_subjects = n_distinct(sub),
    total_trials = n(),
    total_ch2_ready = if ("gate_pupil_primary" %in% names(ch2_data)) {
      sum(gate_pupil_primary, na.rm = TRUE)
    } else if ("pass_primary_060" %in% names(ch2_data)) {
      sum(pass_primary_060, na.rm = TRUE)
    } else {
      sum(baseline_quality >= 0.60 & cog_quality >= 0.60, na.rm = TRUE)
    },
    total_auc_available = sum(auc_available == TRUE, na.rm = TRUE),
    subjects_with_ch2_data = if ("gate_pupil_primary" %in% names(ch2_data)) {
      n_distinct(sub[gate_pupil_primary == TRUE])
    } else if ("pass_primary_060" %in% names(ch2_data)) {
      n_distinct(sub[pass_primary_060 == TRUE])
    } else {
      n_distinct(sub[baseline_quality >= 0.60 & cog_quality >= 0.60])
    },
    .groups = "drop"
  ) %>%
  mutate(
    retention_rate = ifelse(total_trials > 0, 
                            100 * total_ch2_ready / total_trials, 0),
    auc_rate = ifelse(total_trials > 0,
                     100 * total_auc_available / total_trials, 0),
    subject_coverage = ifelse(n_subjects > 0,
                             100 * subjects_with_ch2_data / n_subjects, 0)
  )

kable(ch2_summary,
      col.names = c("Task", "N Subjects", "Total Trials", "Ch2 Ready Trials",
                    "AUC Available", "Subjects with Data", "Retention %", "AUC %", "Coverage %"),
      digits = c(0, 0, 0, 0, 0, 0, 1, 1, 1),
      caption = "Chapter 2 Data Availability (60% Threshold)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)

cat("\n**Summary:**\n")
cat("- Total Chapter 2 ready trials: ", format(sum(ch2_summary$total_ch2_ready), big.mark = ","), "\n", sep = "")
cat("- Overall retention rate: ", sprintf("%.1f", mean(ch2_summary$retention_rate)), "%\n", sep = "")
cat("- Subjects with usable data: ", sum(ch2_summary$subjects_with_ch2_data), " / ", sum(ch2_summary$n_subjects), "\n", sep = "")
cat("- Trials with AUC available: ", format(sum(ch2_summary$total_auc_available), big.mark = ","), "\n", sep = "")
```

### 5.3 Per-Participant Chapter 2 Readiness

```{r ch2-participant}
ch2_participant <- ch2_data %>%
  group_by(sub) %>%
  summarise(
    total_trials = n(),
    total_ch2_ready = if ("gate_pupil_primary" %in% names(ch2_data)) {
      sum(gate_pupil_primary, na.rm = TRUE)
    } else if ("pass_primary_060" %in% names(ch2_data)) {
      sum(pass_primary_060, na.rm = TRUE)
    } else {
      sum(baseline_quality >= 0.60 & cog_quality >= 0.60, na.rm = TRUE)
    },
    .groups = "drop"
  ) %>%
  mutate(
    ch2_rate = ifelse(total_trials > 0, 100 * total_ch2_ready / total_trials, 0),
    has_data = total_ch2_ready > 0
  ) %>%
  arrange(desc(total_ch2_ready))

# Show top participants
cat("**Top 10 Participants by Chapter 2 Ready Trials:**\n\n")
kable(head(ch2_participant, 10),
      col.names = c("Subject", "Total Trials", "Ch2 Ready", "Retention %", "Has Data"),
      digits = c(0, 0, 0, 1, 0),
      caption = "Top Participants") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)

cat("\n**Participants with Chapter 2 Data:** ", sum(ch2_participant$has_data), " / ", nrow(ch2_participant), "\n", sep = "")
```

## 6. Chapter 3: Data Readiness

### 6.1 Chapter 3 Requirements

**DDM with Pupil Predictors:** 
- **Quality Gate**: B1 baseline validity ≥ 50% AND Cognitive response window validity ≥ 50%
  - Uses `B1_quality >= 0.50` (not `baseline_quality`) because Cognitive AUC uses B1 baseline for correction
  - Uses `cog_quality >= 0.50` for the cognitive response window
- **Behavioral Filter**: RT between 0.2s and 3.0s (excludes anticipatory and timeout responses)
- **Pupil Metrics**: Cognitive AUC as predictor of drift rate, boundary separation, etc.
- **Note**: The `ddm_ready` flag in the data includes these requirements: `B1_quality >= 0.50` AND `cog_quality >= 0.50` AND RT 0.2-3.0s

**Behavior-Only DDM:**

-   **No pupil requirements**: Uses all behavioral trials with valid RT
-   **Larger sample size**: Can include trials with poor or missing
    pupil data

### 6.2 Chapter 3 Data Availability

```{r ch3-availability}
ch3_summary <- ch3_data %>%
  group_by(task) %>%
  summarise(
    n_subjects = n_distinct(sub),
    total_trials = n(),
    total_ddm_ready = if ("ddm_ready" %in% names(ch3_data)) {
      sum(ddm_ready, na.rm = TRUE)
    } else {
      # Fallback: use B1_quality if available, otherwise baseline_quality
      if ("B1_quality" %in% names(ch3_data)) {
        sum(B1_quality >= 0.50 & cog_quality >= 0.50 & 
            !is.na(rt) & rt >= 0.2 & rt <= 3.0, na.rm = TRUE)
      } else {
        sum(baseline_quality >= 0.50 & cog_quality >= 0.50 & 
            !is.na(rt) & rt >= 0.2 & rt <= 3.0, na.rm = TRUE)
      }
    },
    total_auc_available = sum(auc_available == TRUE, na.rm = TRUE),
    subjects_with_ch3_data = if ("ddm_ready" %in% names(ch3_data)) {
      n_distinct(sub[ddm_ready == TRUE])
    } else {
      # Fallback: use B1_quality if available
      if ("B1_quality" %in% names(ch3_data)) {
        n_distinct(sub[B1_quality >= 0.50 & cog_quality >= 0.50 & 
                       !is.na(rt) & rt >= 0.2 & rt <= 3.0])
      } else {
        n_distinct(sub[baseline_quality >= 0.50 & cog_quality >= 0.50 & 
                       !is.na(rt) & rt >= 0.2 & rt <= 3.0])
      }
    },
    .groups = "drop"
  ) %>%
  mutate(
    ddm_retention = ifelse(total_trials > 0,
                           100 * total_ddm_ready / total_trials, 0),
    auc_rate = ifelse(total_trials > 0,
                     100 * total_auc_available / total_trials, 0),
    subject_coverage = ifelse(n_subjects > 0,
                              100 * subjects_with_ch3_data / n_subjects, 0)
  )

kable(ch3_summary,
      col.names = c("Task", "N Subjects", "Total Trials", "DDM Ready",
                    "AUC Available", "Subjects with Data", "DDM %", "AUC %", "Coverage %"),
      digits = c(0, 0, 0, 0, 0, 0, 1, 1, 1),
      caption = "Chapter 3 Data Availability (50% Threshold + RT Filter)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)

# Calculate intersection: ddm_ready & auc_available (required for pupil-enhanced DDM)
ch3_intersection <- ch3_data %>%
  filter(if ("ddm_ready" %in% names(ch3_data)) ddm_ready == TRUE else {
    if ("B1_quality" %in% names(ch3_data)) {
      B1_quality >= 0.50 & cog_quality >= 0.50 & !is.na(rt) & rt >= 0.2 & rt <= 3.0
    } else {
      baseline_quality >= 0.50 & cog_quality >= 0.50 & !is.na(rt) & rt >= 0.2 & rt <= 3.0
    }
  }) %>%
  filter(auc_available == TRUE) %>%
  group_by(task) %>%
  summarise(n_intersection = n(), .groups = "drop")

cat("\n**Summary:**\n")
cat("- Total Chapter 3 ready trials (DDM-ready): ", format(sum(ch3_summary$total_ddm_ready), big.mark = ","), "\n", sep = "")
cat("- Overall DDM retention rate: ", sprintf("%.1f", mean(ch3_summary$ddm_retention)), "%\n", sep = "")
cat("- Subjects with usable data: ", sum(ch3_summary$subjects_with_ch3_data), " / ", sum(ch3_summary$n_subjects), "\n", sep = "")
cat("- Trials with AUC available: ", format(sum(ch3_summary$total_auc_available), big.mark = ","), "\n", sep = "")
cat("\n**For pupil-enhanced DDM, the final analysis set is `ddm_ready & auc_available`:**\n")
cat("- Total trials meeting both criteria: ", format(sum(ch3_intersection$n_intersection), big.mark = ","), "\n", sep = "")
cat("  - ADT: ", format(ch3_intersection$n_intersection[ch3_intersection$task == "ADT"], big.mark = ","), " trials\n", sep = "")
cat("  - VDT: ", format(ch3_intersection$n_intersection[ch3_intersection$task == "VDT"], big.mark = ","), " trials\n", sep = "")
```

### 6.3 Behavior-Only vs Pupil-Enhanced DDM

```{r ch3-comparison}
# Estimate behavior-only from join health
behavior_only_trials <- sum(join_health$n_behavioral, na.rm = TRUE)

ch3_comparison <- data.frame(
  Analysis_Type = c("Behavior-Only DDM", "DDM with Pupil Predictors"),
  Total_Trials = c(
    behavior_only_trials,
    sum(ch3_summary$total_ddm_ready, na.rm = TRUE)
  ),
  N_Subjects = c(
    length(unique(join_health$sub)),
    sum(ch3_summary$subjects_with_ch3_data, na.rm = TRUE)
  )
) %>%
  mutate(
    Advantage = case_when(
      Analysis_Type == "Behavior-Only DDM" ~ "Larger sample, no pupil quality requirements",
      Analysis_Type == "DDM with Pupil Predictors" ~ "Can test arousal effects on decision-making"
    )
  )

kable(ch3_comparison,
      col.names = c("Analysis Type", "Total Trials", "N Subjects", "Advantage"),
      caption = "Chapter 3 Analysis Options") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## 7. Data Quality Issues and Considerations

### 7.1 Common Data Quality Challenges

1.  **Missing Samples**: Blinks, eye movements, or tracking loss result
    in missing pupil diameter measurements
2.  **Window Validity**: Critical windows (baseline, cognitive) must
    have sufficient valid samples for reliable metrics
3.  **Behavioral Alignment**: RT data must be properly aligned with
    pupil time series
4.  **Trial Exclusions**: Trials with window out-of-bounds or all-NaN
    data are excluded

### 7.2 Quality Control Measures

-   **Pre-filtering**: Excludes trials with `window_oob==1` or
    `all_nan==1`
-   **Validity Thresholds**: Multiple thresholds (40%, 50%, 60%, 70%)
    allow sensitivity analyses
-   **Gate System**: Independent gates for different analysis types
    (baseline, cognitive, overall)

### 7.3 Recommendations

Based on the data availability:

1.  **For Chapter 2**:

    -   Primary analysis at 60% threshold provides conservative,
        high-quality data
    -   Sensitivity analyses at 50% and 70% thresholds for robustness
    -   Consider per-participant inclusion based on minimum trial counts

2.  **For Chapter 3**:

    -   DDM with pupil predictors: Use 50% threshold to maximize sample
        size while maintaining data quality
    -   Behavior-only DDM: Can use all behavioral trials, providing
        largest possible sample
    -   Consider running both analyses to compare results with and
        without pupil predictors

3.  **General**:

    -   Monitor window validity distributions to identify systematic
        issues
    -   Consider task-specific thresholds if validity differs
        substantially between ADT and VDT
    -   Document all exclusion criteria and thresholds in methods
        sections

## 8. Participant-Level Data Quality Supplement

This supplement provides detailed visualizations and diagnostics to distinguish between low AUC values due to poor data quality versus genuine low pupil dilation responses.

**Key Concern Addressed**: Low AUC values could result from either (1) poor data quality (many missing samples, large gaps during critical periods) or (2) genuine low pupil dilation responses. This supplement implements best-practice diagnostics to identify which is the case.

**Important Note on AUC Interpretation**: The primary cognitive AUC window is now a **fixed 1.20s stimulus-locked window** (4.85s to 6.05s), which eliminates RT-AUC coupling confounds. However, for response-locked windows (secondary analyses), raw AUC values are mechanically tied to response time (RT) because longer RT = more time to accumulate area, even if pupil response amplitude is identical. Therefore, we compute **RT-normalized metrics** (`cog_mean = cog_auc / window_duration`) to separate amplitude effects from duration effects. **For primary analyses, prefer `cog_mean` over raw `cog_auc` for interpretability.**

### Mean Dilation Metrics and Quality Diagnostics

```{r rt-normalized-metrics}
# Compute mean dilation (AUC / window_duration) for cognitive window
# For primary window (fixed 1.20s), this removes any residual duration effects
# For response-locked windows, this removes RT-AUC coupling
ch2_data_enhanced <- ch2_data %>%
  mutate(
    # Primary window is fixed: 4.85s to 6.05s (1.20s duration)
    # cog_window_duration should already be computed in the data
    # If missing, compute from window definition
    cog_window_duration = if_else(
      is.na(cog_window_duration),
      1.20,  # Fixed primary window duration
      cog_window_duration
    ),
    
    # Mean dilation (AUC / window_duration) - preferred metric
    cog_mean = if_else(!is.na(cog_auc) & !is.na(cog_window_duration) & cog_window_duration > 0, 
                       cog_auc / cog_window_duration, 
                       NA_real_),
    
    # Total window duration for normalization (if needed)
    total_window_duration = cog_window_end - 0,  # From squeeze onset
    total_mean = if_else(!is.na(total_auc) & total_window_duration > 0,
                         total_auc / total_window_duration,
                         NA_real_),
    
    # RT bin for visualization
    rt_bin = case_when(
      is.na(rt) ~ "No RT",
      rt < 0.5 ~ "< 0.5s",
      rt < 1.0 ~ "0.5-1.0s",
      rt < 1.5 ~ "1.0-1.5s",
      rt < 2.0 ~ "1.5-2.0s",
      TRUE ~ "≥ 2.0s"
    )
  )

# Key diagnostic: Scatter plot of RT-normalized pupil metric vs cognitive quality
# This directly addresses "low AUC due to quality vs low AUC due to low dilation"
diagnostic_data <- ch2_data_enhanced %>%
  filter(!is.na(cog_mean) & !is.na(cog_quality)) %>%
  mutate(
    tooltip_text = paste0(
      "Participant: ", sub, "<br>",
      "Task: ", task, "<br>",
      "Trial: ", trial_index, "<br>",
      "Cognitive Quality: ", round(cog_quality, 3), "<br>",
      "RT-Normalized AUC: ", round(cog_mean, 3), "<br>",
      "Window Duration: ", round(cog_window_duration, 2), "s"
    )
  )

diagnostic_scatter <- diagnostic_data %>%
  ggplot(aes(x = cog_quality, y = cog_mean, color = task, size = cog_window_duration,
             text = tooltip_text)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
  facet_wrap(~ task, ncol = 2) +
  labs(
    x = "Cognitive Quality (proportion valid samples)",
    y = "RT-Normalized Cognitive AUC (mean pupil in window)",
    color = "Task",
    size = "Window\nDuration (s)",
    title = "Diagnostic: RT-Normalized Pupil Response vs Data Quality",
    subtitle = "Hover over points to see participant IDs. Low values at HIGH quality = genuine low dilation. Low values at LOW quality = data quality issue."
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(face = "bold")
  )

# Convert to plotly for interactivity (hover to see participant IDs)
diagnostic_scatter_ly <- ggplotly(diagnostic_scatter, tooltip = "text") %>%
  layout(hovermode = "closest")

print(diagnostic_scatter_ly)

# Summary statistics
cat("\n**Key Diagnostic Interpretation:**\n\n")
cat("- **Points in upper-right quadrant** (high quality, high dilation): Normal responses with good data\n")
cat("- **Points in lower-right quadrant** (high quality, low dilation): Genuine low dilation responses\n")
cat("- **Points in lower-left quadrant** (low quality, low dilation): Likely data quality artifacts\n")
cat("- **Slope of regression line**: If positive, suggests quality-AUC relationship (data quality issue)\n")
cat("- **If slope is flat**: Quality and AUC are independent (low AUC is physiological, not artifactual)\n\n")

# Correlation between quality and RT-normalized AUC
quality_auc_cor <- ch2_data_enhanced %>%
  filter(!is.na(cog_mean) & !is.na(cog_quality)) %>%
  group_by(task) %>%
  summarise(
    cor_quality_auc = cor(cog_quality, cog_mean, use = "complete.obs"),
    n_trials = n(),
    .groups = "drop"
  )

cat("**Correlation between Cognitive Quality and RT-Normalized AUC:**\n\n")
kable(quality_auc_cor,
      col.names = c("Task", "Correlation", "N Trials"),
      caption = "Quality-AUC Relationship",
      digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

### Participant-Task Combinations (Using RT-Normalized Metrics)

```{r participant-supplement, fig.height=8, fig.width=12}
# Prepare participant-level data using RT-normalized metrics
# Use the enhanced data with RT-normalized cognitive AUC
participant_auc_data <- ch2_data_enhanced %>%
  filter(!is.na(sub) & !is.na(task)) %>%
  group_by(sub, task) %>%
  summarise(
    n_trials = n(),
    n_with_total_auc = sum(!is.na(total_auc), na.rm = TRUE),
    n_with_cog_auc = sum(!is.na(cog_auc), na.rm = TRUE),
    mean_baseline_quality = mean(baseline_quality, na.rm = TRUE),
    mean_cog_quality = mean(cog_quality, na.rm = TRUE),
    mean_total_auc = mean(total_auc, na.rm = TRUE),
    mean_cog_auc = mean(cog_auc, na.rm = TRUE),
    mean_cog_mean = mean(cog_mean, na.rm = TRUE),  # RT-normalized
    median_total_auc = median(total_auc, na.rm = TRUE),
    median_cog_auc = median(cog_auc, na.rm = TRUE),
    median_cog_mean = median(cog_mean, na.rm = TRUE),  # RT-normalized
    sd_total_auc = sd(total_auc, na.rm = TRUE),
    sd_cog_auc = sd(cog_auc, na.rm = TRUE),
    sd_cog_mean = sd(cog_mean, na.rm = TRUE),  # RT-normalized
    pct_trials_with_auc = 100 * sum(!is.na(total_auc) & !is.na(cog_auc), na.rm = TRUE) / n(),
    .groups = "drop"
  ) %>%
  mutate(
    data_quality_label = case_when(
      mean_baseline_quality >= 0.7 & mean_cog_quality >= 0.7 ~ "High Quality",
      mean_baseline_quality >= 0.5 & mean_cog_quality >= 0.5 ~ "Moderate Quality",
      TRUE ~ "Low Quality"
    )
  )

# Get unique participant-task combinations
participant_tasks <- participant_auc_data %>%
  select(sub, task) %>%
  distinct() %>%
  arrange(sub, task)

cat("**Total participant-task combinations:**", nrow(participant_tasks), "\n\n")

# Summary plot: RT-normalized cognitive AUC by participant-task
# This removes RT effects and focuses on amplitude
summary_plot <- participant_auc_data %>%
  ggplot(aes(x = reorder(paste(sub, task, sep = " - "), mean_cog_mean), 
             y = mean_cog_mean, fill = data_quality_label)) +
  geom_col(alpha = 0.8) +
  facet_wrap(~ task, scales = "free_x", ncol = 1) +
  labs(
    x = "Participant - Task",
    y = "Mean RT-Normalized Cognitive AUC\n(Mean Pupil in Cognitive Window)",
    fill = "Data Quality",
    title = "RT-Normalized Cognitive AUC by Participant and Task",
    subtitle = "Removes RT effects. Low values at HIGH quality = genuine low dilation. Low values at LOW quality = data quality issue."
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    legend.position = "bottom",
    plot.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(face = "bold")
  )

print(summary_plot)

# Identify problematic cases (low quality OR low AUC with high quality - needs investigation)
problematic_cases <- participant_auc_data %>%
  mutate(
    problem_score = case_when(
      data_quality_label == "Low Quality" ~ 3,  # Most problematic
      data_quality_label == "Moderate Quality" & mean_cog_mean < quantile(participant_auc_data$mean_cog_mean, 0.25, na.rm = TRUE) ~ 2,
      mean_cog_mean < quantile(participant_auc_data$mean_cog_mean, 0.1, na.rm = TRUE) & mean_cog_quality > 0.6 ~ 1,  # Low AUC but high quality (needs check)
      TRUE ~ 0
    )
  ) %>%
  filter(problem_score > 0) %>%
  arrange(desc(problem_score), mean_cog_quality) %>%
  head(5)

# Identify good exemplars (high quality, normal AUC)
good_exemplars <- participant_auc_data %>%
  filter(data_quality_label == "High Quality",
         mean_cog_mean > quantile(participant_auc_data$mean_cog_mean, 0.5, na.rm = TRUE),
         mean_cog_mean < quantile(participant_auc_data$mean_cog_mean, 0.9, na.rm = TRUE)) %>%
  arrange(mean_cog_mean) %>%
  head(2)

cat("\n\n**Detailed Examination: Top 5 Most Problematic Cases + 2 Good Exemplars**\n\n")
cat("**Note:** Full participant-level plots for all combinations are available in the complete supplement.\n\n")

# Combine problematic and good cases
example_participants <- bind_rows(
  problematic_cases %>% mutate(case_type = "Problematic"),
  good_exemplars %>% mutate(case_type = "Good Exemplar")
) %>%
  select(sub, task, case_type)

for (i in 1:nrow(example_participants)) {
  p_sub <- example_participants$sub[i]
  p_task <- example_participants$task[i]
  
  # Get trial-level data for this participant-task (using enhanced data with RT-normalized metrics)
  p_data <- ch2_data_enhanced %>%
    filter(sub == p_sub, task == p_task) %>%
    select(trial_index, total_auc, cog_auc, cog_mean, baseline_quality, cog_quality, 
           cog_window_duration, rt, auc_available, auc_missing_reason) %>%
    mutate(
      trial_num = row_number(),
      has_both_auc = !is.na(total_auc) & !is.na(cog_auc),
      has_cog_mean = !is.na(cog_mean)
    )
  
  if (nrow(p_data) == 0) next
  
  # Create barplot for RT-normalized cognitive AUC (preferred over raw AUC)
  if (sum(p_data$has_cog_mean) > 0) {
    p_auc_plot <- p_data %>%
      filter(has_cog_mean) %>%
      ggplot(aes(x = trial_num, y = cog_mean, fill = cog_quality)) +
      geom_col(alpha = 0.8, width = 0.8) +
      scale_fill_gradient2(low = "red", mid = "yellow", high = "green", 
                          midpoint = 0.6, name = "Cog Quality") +
      labs(
        x = "Trial Number",
        y = "RT-Normalized Cognitive AUC\n(Mean Pupil in Window)",
        fill = "Cog Quality",
        title = paste0("Participant ", p_sub, " - ", p_task, ": RT-Normalized Cognitive AUC by Trial"),
        subtitle = paste0("Trials with AUC: ", sum(p_data$has_cog_mean), " / ", nrow(p_data), 
                         " (", round(100 * sum(p_data$has_cog_mean) / nrow(p_data), 1), "%)",
                         " | Color = Cognitive Quality")
      ) +
      theme_minimal() +
      theme(
        legend.position = "bottom",
        plot.title = element_text(size = 11, face = "bold")
      )
    
    # Quality summary
    quality_summary <- participant_auc_data %>%
      filter(sub == p_sub, task == p_task)
    
    if (nrow(quality_summary) > 0) {
      cat("\n\n#### ", p_sub, " - ", p_task, "\n\n", sep = "")
      cat("**Data Quality Summary:**\n")
      cat("- Total Trials:", quality_summary$n_trials, "\n")
      cat("- Trials with Total AUC:", quality_summary$n_with_total_auc, "\n")
      cat("- Trials with Cognitive AUC:", quality_summary$n_with_cog_auc, "\n")
      cat("- Mean Baseline Quality:", round(quality_summary$mean_baseline_quality, 3), "\n")
      cat("- Mean Cognitive Quality:", round(quality_summary$mean_cog_quality, 3), "\n")
      cat("- Mean Total AUC:", round(quality_summary$mean_total_auc, 2), "\n")
      cat("- Mean Cognitive AUC (raw):", round(quality_summary$mean_cog_auc, 2), "\n")
      cat("- Mean RT-Normalized Cognitive AUC:", round(quality_summary$mean_cog_mean, 3), "\n")
      cat("- Data Quality Label:", quality_summary$data_quality_label, "\n\n")
      
      print(p_auc_plot)
    }
  } else {
    cat("\n\n#### ", p_sub, " - ", p_task, "\n\n", sep = "")
    cat("**No AUC data available for this participant-task combination.**\n")
    cat("This may indicate data quality issues.\n\n")
  }
}

cat("\n\n**Note:** This section shows the most critical cases for advisor review. Full participant-level plots for all ", nrow(participant_tasks), " combinations are available in a separate supplement document or can be generated on request.\n", sep = "")
```

### Gap-Aware AUC Quality Control Metrics

**Rationale**: Percentage-validity thresholds (50-60%) are necessary but not sufficient. A trial can have 60% valid samples overall, but if there's a **large contiguous gap** (e.g., 400ms) during the peak response period, the trapezoidal AUC will bridge across that gap with a straight line, potentially underestimating or distorting the true pupil response. This section implements gap-aware QC metrics recommended by Burg et al. and Kret & Sjak-Shie.

**New QC Metrics Computed:**

1. **`cog_auc_max_gap_ms`**: Maximum contiguous missing segment (in milliseconds) within the cognitive AUC window
2. **`cog_auc_n_segments`**: Number of contiguous valid segments (more segments = more fragmented sampling)
3. **`cog_auc_n_valid`**: Count of valid (non-NA) samples in the cognitive window
4. **`cog_window_duration`**: Duration of the cognitive window (should be ~1.20s for fixed primary window: 4.85s to 6.05s)
5. **`cog_auc_prop_valid`**: Proportion of valid samples (n_valid / expected_samples_at_250Hz)

**Gap-Aware AUC Calculation**: The AUC function now **does not bridge across gaps larger than 250ms**. Instead, it:
- Splits the window into contiguous segments where time differences ≤ 250ms
- Computes trapezoidal AUC within each segment separately
- Sums segment AUCs to get total AUC
- This ensures AUC reflects "area under observed signal where we had continuous coverage"

```{r gap-aware-prepare-data, warning=FALSE, message=FALSE}
# Check if gap-aware metrics are available (they should be if data was processed with updated pipeline)
has_gap_metrics <- all(c("cog_auc_max_gap_ms", "cog_auc_n_segments", "cog_auc_n_valid", 
                          "cog_window_duration", "cog_auc_prop_valid") %in% names(ch2_data))

if (has_gap_metrics) {
  # Enhanced data with gap-aware metrics
  ch2_data_gap_enhanced <- ch2_data %>%
    mutate(
      # RT-normalized cognitive AUC (if not already computed)
      cog_mean = if_else(!is.na(cog_auc) & !is.na(cog_window_duration) & cog_window_duration > 0,
                         cog_auc / cog_window_duration, NA_real_),
      # Gap severity categories
      gap_severity = case_when(
        is.na(cog_auc_max_gap_ms) ~ "No Data",
        cog_auc_max_gap_ms <= 150 ~ "Small Gap (≤150ms)",
        cog_auc_max_gap_ms <= 250 ~ "Moderate Gap (150-250ms)",
        cog_auc_max_gap_ms <= 400 ~ "Large Gap (250-400ms)",
        TRUE ~ "Very Large Gap (>400ms)"
      ),
      # Segment fragmentation
      fragmentation = case_when(
        is.na(cog_auc_n_segments) ~ "No Data",
        cog_auc_n_segments == 1 ~ "Continuous",
        cog_auc_n_segments <= 3 ~ "Mildly Fragmented",
        cog_auc_n_segments <= 5 ~ "Moderately Fragmented",
        TRUE ~ "Highly Fragmented"
      ),
      # Window duration categories (for short RT concerns)
      window_duration_cat = case_when(
        is.na(cog_window_duration) ~ "No Data",
        cog_window_duration < 0.3 ~ "Very Short (<0.3s)",
        cog_window_duration < 0.5 ~ "Short (0.3-0.5s)",
        cog_window_duration < 1.0 ~ "Moderate (0.5-1.0s)",
        TRUE ~ "Long (≥1.0s)"
      )
    )
} else {
  cat("\n**Gap-aware QC metrics not yet available in current dataset.**\n")
  cat("These metrics will be computed when the data pipeline is re-run with the updated `calculate_auc_with_qc()` function.\n")
}
```

### Diagnostic 1: RT-Normalized AUC vs Quality, Colored by Maximum Gap

```{r gap-aware-diagnostic-1, fig.height=8, fig.width=12, warning=FALSE, message=FALSE}
if (exists("has_gap_metrics") && has_gap_metrics && exists("ch2_data_gap_enhanced")) {
  diagnostic_gap_scatter <- ch2_data_gap_enhanced %>%
    filter(!is.na(cog_mean) & !is.na(cog_quality) & !is.na(cog_auc_max_gap_ms)) %>%
    ggplot(aes(x = cog_quality, y = cog_mean, color = gap_severity, 
               size = cog_window_duration,
               text = paste0("Participant: ", sub, "<br>",
                            "Task: ", task, "<br>",
                            "Cog Quality: ", round(cog_quality, 3), "<br>",
                            "RT-Normalized AUC: ", round(cog_mean, 3), "<br>",
                            "Max Gap: ", round(cog_auc_max_gap_ms, 1), "ms", "<br>",
                            "Window Duration: ", round(cog_window_duration, 2), "s", "<br>",
                            "N Segments: ", cog_auc_n_segments))) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed", size = 0.5) +
    facet_wrap(~ task, ncol = 2) +
    scale_color_manual(values = c("Small Gap (≤150ms)" = "green", 
                                   "Moderate Gap (150-250ms)" = "yellow",
                                   "Large Gap (250-400ms)" = "orange",
                                   "Very Large Gap (>400ms)" = "red",
                                   "No Data" = "gray")) +
    labs(
      x = "Cognitive Quality (proportion valid samples)",
      y = "RT-Normalized Cognitive AUC (mean pupil in window)",
      color = "Max Gap",
      size = "Window\nDuration (s)",
      title = "Diagnostic: RT-Normalized AUC vs Quality, Colored by Maximum Gap",
      subtitle = "Large gaps (>250ms) can distort AUC even when overall quality looks acceptable"
    ) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(size = 12, face = "bold"),
      strip.text = element_text(face = "bold")
    )
  
  ggplotly(diagnostic_gap_scatter, tooltip = "text")
}
```

### Diagnostic 2: Distribution of Maximum Gaps by Quality Tier

```{r gap-aware-diagnostic-2, fig.height=8, fig.width=12, warning=FALSE, message=FALSE}
if (exists("has_gap_metrics") && has_gap_metrics && exists("ch2_data_gap_enhanced")) {
  gap_distribution <- ch2_data_gap_enhanced %>%
    filter(!is.na(cog_auc_max_gap_ms) & !is.na(cog_quality)) %>%
    mutate(
      quality_tier = case_when(
        cog_quality >= 0.70 ~ "High (≥70%)",
        cog_quality >= 0.60 ~ "Moderate-High (60-70%)",
        cog_quality >= 0.50 ~ "Moderate (50-60%)",
        TRUE ~ "Low (<50%)"
      )
    ) %>%
    ggplot(aes(x = cog_auc_max_gap_ms, fill = quality_tier)) +
    geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
    geom_vline(xintercept = 250, linetype = "dashed", color = "red", size = 1) +
    facet_wrap(~ task, ncol = 2) +
    labs(
      x = "Maximum Gap (ms)",
      y = "Number of Trials",
      fill = "Quality Tier",
      title = "Distribution of Maximum Gaps by Quality Tier",
      subtitle = "Red dashed line = 250ms threshold (Kret & Sjak-Shie recommendation)"
    ) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(size = 12, face = "bold"),
      strip.text = element_text(face = "bold")
    )
  
  print(gap_distribution)
}
```

### Diagnostic 3: RT-Normalized AUC vs Window Duration

```{r gap-aware-diagnostic-3, fig.height=8, fig.width=12, warning=FALSE, message=FALSE}
if (exists("has_gap_metrics") && has_gap_metrics && exists("ch2_data_gap_enhanced")) {
  duration_scatter <- ch2_data_gap_enhanced %>%
    filter(!is.na(cog_mean) & !is.na(cog_window_duration) & !is.na(cog_quality)) %>%
    ggplot(aes(x = cog_window_duration, y = cog_mean, color = cog_quality,
               text = paste0("Participant: ", sub, "<br>",
                            "Task: ", task, "<br>",
                            "Window Duration: ", round(cog_window_duration, 2), "s", "<br>",
                            "RT-Normalized AUC: ", round(cog_mean, 3), "<br>",
                            "Cog Quality: ", round(cog_quality, 3)))) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed", size = 0.5) +
    facet_wrap(~ task, ncol = 2) +
    scale_color_gradient2(low = "red", mid = "yellow", high = "green", midpoint = 0.6) +
    labs(
      x = "Cognitive Window Duration (s)",
      y = "RT-Normalized Cognitive AUC (mean pupil in window)",
      color = "Cog Quality",
      title = "Diagnostic: RT-Normalized AUC vs Window Duration",
      subtitle = "Short windows (fast RTs) may yield less stable estimates even with high quality"
    ) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(size = 12, face = "bold"),
      strip.text = element_text(face = "bold")
    )
  
  ggplotly(duration_scatter, tooltip = "text")
}
```

```{r gap-aware-summary-table, warning=FALSE, message=FALSE}
if (exists("has_gap_metrics") && has_gap_metrics && exists("ch2_data_gap_enhanced")) {
  # Summary table: Retention rates with gap-aware exclusions
  # Use trials that have gap metrics available (don't filter out NA gap metrics - they're valid "no data" cases)
  gap_exclusion_summary <- ch2_data_gap_enhanced %>%
    filter(!is.na(cog_quality)) %>%  # Only require quality to be non-NA
    group_by(task) %>%
    summarise(
      n_total = n(),
      # Standard 60% threshold
      n_pass_60 = sum(baseline_quality >= 0.60 & cog_quality >= 0.60, na.rm = TRUE),
      # 60% + max_gap <= 250ms (only count if gap metric exists)
      n_pass_60_gap250 = sum(baseline_quality >= 0.60 & cog_quality >= 0.60 & 
                             !is.na(cog_auc_max_gap_ms) & cog_auc_max_gap_ms <= 250, na.rm = TRUE),
      # 60% + max_gap <= 250ms + min duration >= 0.5s (only count if both metrics exist)
      n_pass_60_gap250_dur50 = sum(baseline_quality >= 0.60 & cog_quality >= 0.60 & 
                                    !is.na(cog_auc_max_gap_ms) & cog_auc_max_gap_ms <= 250 &
                                    !is.na(cog_window_duration) & cog_window_duration >= 0.5, na.rm = TRUE),
      # 60% + max_gap <= 250ms + min duration >= 0.5s + n_valid >= 100 (only count if all metrics exist)
      n_pass_60_gap250_dur50_n100 = sum(baseline_quality >= 0.60 & cog_quality >= 0.60 & 
                                         !is.na(cog_auc_max_gap_ms) & cog_auc_max_gap_ms <= 250 &
                                         !is.na(cog_window_duration) & cog_window_duration >= 0.5 &
                                         !is.na(cog_auc_n_valid) & cog_auc_n_valid >= 100, na.rm = TRUE),
      # Count how many trials have gap metrics available
      n_with_gap_metrics = sum(!is.na(cog_auc_max_gap_ms), na.rm = TRUE),
      n_with_duration = sum(!is.na(cog_window_duration), na.rm = TRUE),
      n_with_n_valid = sum(!is.na(cog_auc_n_valid), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      pct_pass_60 = 100 * n_pass_60 / n_total,
      pct_pass_60_gap250 = 100 * n_pass_60_gap250 / n_total,
      pct_pass_60_gap250_dur50 = 100 * n_pass_60_gap250_dur50 / n_total,
      pct_pass_60_gap250_dur50_n100 = 100 * n_pass_60_gap250_dur50_n100 / n_total
    )
  
  # Check if we have meaningful data
  if (sum(gap_exclusion_summary$n_with_gap_metrics) == 0) {
    cat("\n**Note:** Gap-aware metrics (`cog_auc_max_gap_ms`, `cog_window_duration`, `cog_auc_n_valid`) are not available in the current dataset.\n")
    cat("The diagnostic plots above are shown for conceptual illustration. Once the pipeline is re-run with gap-aware AUC calculation,\n")
    cat("these metrics will be computed and this table will show meaningful retention rates.\n\n")
  } else {
    cat("\n**Retention Rates with Gap-Aware Exclusions (Chapter 2, 60% Base Threshold):**\n")
    cat("**Note:** Rows with duration and N-valid criteria only count trials where those metrics are available.\n")
    cat("Trials with missing gap metrics are excluded from those specific criteria but included in the '60% Only' baseline.\n\n")
    
    kable(gap_exclusion_summary %>%
          select(task, n_total, n_pass_60, pct_pass_60, 
                 n_pass_60_gap250, pct_pass_60_gap250,
                 n_pass_60_gap250_dur50, pct_pass_60_gap250_dur50,
                 n_pass_60_gap250_dur50_n100, pct_pass_60_gap250_dur50_n100) %>%
          mutate(across(where(is.numeric), ~ round(.x, 1))),
          col.names = c("Task", "Total", "60% Only", "60% %", 
                        "60%+Gap≤250", "60%+Gap %",
                        "60%+Gap+Dur≥0.5s", "60%+Gap+Dur %",
                        "60%+Gap+Dur+N≥100", "60%+Gap+Dur+N %"),
          caption = "Sensitivity: How Gap-Aware Exclusions Affect Retention") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                    full_width = FALSE)
    
    # Diagnostic: Show how many trials have gap metrics
    cat("\n**Gap Metric Availability:**\n")
    cat("- Trials with `cog_auc_max_gap_ms`: ", sum(gap_exclusion_summary$n_with_gap_metrics), " / ", sum(gap_exclusion_summary$n_total), "\n", sep = "")
    cat("- Trials with `cog_window_duration`: ", sum(gap_exclusion_summary$n_with_duration), " / ", sum(gap_exclusion_summary$n_total), "\n", sep = "")
    cat("- Trials with `cog_auc_n_valid`: ", sum(gap_exclusion_summary$n_with_n_valid), " / ", sum(gap_exclusion_summary$n_total), "\n\n", sep = "")
  }
}
```

### Interpreting the Gap-Aware Diagnostic Plots

**Purpose**: These diagnostic plots help distinguish between **genuine low pupil dilation** (valid physiological data) and **data quality artifacts** (missing data during critical periods). This is essential for making informed decisions about which trials to include in analyses.

#### Diagnostic 1: RT-Normalized AUC vs Quality, Colored by Maximum Gap

**What This Plot Shows:**
- **X-axis**: Cognitive Quality (proportion of valid samples in the cognitive window)
- **Y-axis**: RT-Normalized Cognitive AUC (mean pupil dilation, removing RT effects)
- **Color**: Maximum gap size within the cognitive window
- **Size**: Window duration (larger points = longer RTs)

**How to Interpret:**

1. **Lower-Right Quadrant (High Quality, Low Dilation, Small Gaps) = ✅ KEEP**
   - **Pattern**: Points with `cog_quality ≥ 0.6`, low `cog_mean`, and green/yellow colors (gaps ≤ 250ms)
   - **Meaning**: These represent genuine low arousal/dilation responses with good data coverage
   - **Action**: **Include in analyses** - these are valid physiological data

2. **Lower-Left Quadrant (Low Quality OR Large Gaps, Low Dilation) = ❌ EXCLUDE**
   - **Pattern**: Points with low `cog_quality` (< 0.5) OR red/orange colors (gaps > 250ms), and low `cog_mean`
   - **Meaning**: Low AUC is likely due to missing data during critical periods, not genuine low dilation
   - **Action**: **Exclude from primary analyses** - these are data quality artifacts

3. **Regression Slope Interpretation:**
   - **Flat/Negative Slope**: Quality and AUC are independent → Good sign! Low AUC values are physiological, not artifactual
   - **Positive Slope**: Suggests quality-AUC relationship → Data quality issue; consider stricter exclusions

4. **Color Distribution:**
   - **Mostly Green/Yellow**: Good data coverage with small gaps → High confidence in AUC estimates
   - **Many Red/Orange Points**: Large gaps present → Consider gap-aware exclusions

**Decision Rule**: If a trial has low RT-normalized AUC, check its color:
- **Green/Yellow (small gap) + High Quality**: Keep (genuine low dilation)
- **Red/Orange (large gap) OR Low Quality**: Exclude (data quality artifact)

#### Diagnostic 2: Distribution of Maximum Gaps by Quality Tier

**What This Plot Shows:**
- **X-axis**: Maximum gap size (ms) within the cognitive window
- **Y-axis**: Number of trials
- **Color**: Quality tier (High ≥70%, Moderate-High 60-70%, Moderate 50-60%, Low <50%)
- **Red dashed line**: 250ms threshold (Kret & Sjak-Shie recommendation)

**How to Interpret:**

1. **Gap Distribution Patterns:**
   - **Left-skewed (most gaps < 250ms)**: Good data coverage → High confidence in AUC estimates
   - **Right-skewed (many gaps > 250ms)**: Problematic → Large gaps can distort AUC even with acceptable %-valid

2. **Quality Tier vs Gap Size:**
   - **High Quality + Large Gaps**: These trials have good overall coverage but large contiguous missing segments → **Exclude** (gap-aware exclusion)
   - **Low Quality + Large Gaps**: Multiple data quality issues → **Definitely exclude**

3. **Threshold Assessment:**
   - **Most trials < 250ms**: Gap-aware exclusions will have minimal impact on sample size
   - **Many trials > 250ms**: Gap-aware exclusions will reduce sample size but improve data quality

**Decision Rule**: 
- **Trials with gaps > 250ms**: Exclude from primary analyses (following Kret & Sjak-Shie recommendation)
- **Trials with gaps ≤ 250ms**: Include if they also meet quality thresholds

#### Diagnostic 3: RT-Normalized AUC vs Window Duration

**What This Plot Shows:**
- **X-axis**: Cognitive window duration (seconds) - determined by RT
- **Y-axis**: RT-Normalized Cognitive AUC (mean pupil dilation)
- **Color**: Cognitive quality (red = low, green = high)
- **Size**: Not used in this plot

**How to Interpret:**

1. **Window Duration Effects:**
   - **Short Windows (< 0.5s)**: Fast RTs → Less stable AUC estimates even with high quality
   - **Long Windows (≥ 1.0s)**: Slow RTs → More stable AUC estimates

2. **Quality-Color Patterns:**
   - **Green points (high quality) across all durations**: Good data regardless of RT
   - **Red points (low quality) at short durations**: Data quality issues + unstable estimates → **Exclude**

3. **Duration-AUC Relationship:**
   - **Flat relationship**: Window duration doesn't affect RT-normalized AUC → Good (RT normalization working)
   - **Positive relationship**: Suggests duration effects remain → May need stricter duration thresholds

**Decision Rule**:
- **Short windows (< 0.3s) + Low Quality**: Exclude (unstable estimates)
- **Short windows (< 0.3s) + High Quality**: Consider excluding or flag for sensitivity analysis
- **Windows ≥ 0.5s**: Include if they meet quality and gap thresholds

#### Summary Table: Retention Rates with Gap-Aware Exclusions

**What This Table Shows:**
- How many trials pass each exclusion criterion
- Impact of adding gap-aware exclusions on sample size

**How to Interpret:**

1. **60% Only**: Baseline exclusion (quality thresholds only)
2. **60% + Gap ≤ 250ms**: Adds gap-aware exclusion → Shows how many trials have acceptable gaps
3. **60% + Gap + Dur ≥ 0.5s**: Adds minimum window duration → Ensures stable estimates
4. **60% + Gap + Dur + N ≥ 100**: Adds minimum valid samples → Ensures robust estimates

**Decision Making:**
- **Large drop from "60% Only" to "60% + Gap ≤ 250ms"**: Many trials have large gaps → Gap-aware exclusions are important
- **Small drop**: Most trials have acceptable gaps → Gap-aware exclusions have minimal impact
- **Final retention rate**: Use this to plan sample size for analyses

**Recommended Approach:**
1. **Primary Analysis**: Use "60% + Gap + Dur + N ≥ 100" (strictest, highest quality)
2. **Sensitivity Analysis**: Compare with "60% + Gap ≤ 250ms" (moderate) to assess robustness
3. **Report**: Present both retention rates and discuss impact on conclusions

#### Overall Decision Framework

**For Each Trial, Ask:**

1. **Does it meet quality thresholds?** (`baseline_quality ≥ 0.60` AND `cog_quality ≥ 0.60`)
   - ❌ No → Exclude
   - ✅ Yes → Continue to step 2

2. **Does it have acceptable gaps?** (`cog_auc_max_gap_ms ≤ 250`)
   - ❌ No → Exclude (gap-aware exclusion)
   - ✅ Yes → Continue to step 3

3. **Does it have sufficient window duration?** (`cog_window_duration ≥ 0.5s`)
   - ❌ No → Consider excluding or flag for sensitivity analysis
   - ✅ Yes → Continue to step 4

4. **Does it have enough valid samples?** (`cog_auc_n_valid ≥ 100`)
   - ❌ No → Consider excluding or flag for sensitivity analysis
   - ✅ Yes → **Include in primary analysis**

**Key Takeaway**: Low RT-normalized AUC is not necessarily a problem. The diagnostic plots help distinguish:
- **Genuine low dilation** (high quality, small gaps) → **Keep**
- **Data quality artifacts** (low quality OR large gaps) → **Exclude**

**Recommended Exclusion Rules (Based on Literature):**

1. **Primary (Strict)**: `baseline_quality >= 0.60` AND `cog_quality >= 0.60` AND `cog_auc_max_gap_ms <= 250` AND `cog_window_duration >= 0.50` AND `cog_auc_n_valid >= 100`
   - Rationale: Kret & Sjak-Shie recommend not interpolating gaps >250ms; Burg et al. show large gaps can distort AUC even with acceptable %-valid

2. **Sensitivity (Moderate)**: `baseline_quality >= 0.60` AND `cog_quality >= 0.60` AND `cog_auc_max_gap_ms <= 250` AND `cog_window_duration >= 0.30`
   - Rationale: Allows shorter windows (faster RTs) but still enforces gap threshold

3. **Sensitivity (Lenient)**: `baseline_quality >= 0.60` AND `cog_quality >= 0.60` AND `cog_auc_max_gap_ms <= 400`
   - Rationale: Allows larger gaps but still flags very problematic trials

### Interpreting Participant-Level Plots

**Key Diagnostic Framework:**

1. **Use RT-Normalized Metrics**: Always interpret `cog_mean` (RT-normalized cognitive AUC) rather than raw `cog_auc`, because raw AUC is mechanically tied to RT duration.

2. **Low RT-Normalized AUC + High Quality + Small Gap = Genuine Low Dilation**
   - If `cog_mean` is low but `cog_quality` is high (≥0.6) AND `cog_auc_max_gap_ms` is small (≤150ms), this indicates genuine low pupil dilation responses
   - These trials/participants should be **included** in analyses (they represent valid physiological data)

3. **Low RT-Normalized AUC + Low Quality OR Large Gap = Data Quality Artifact**
   - If `cog_mean` is low and (`cog_quality` is low (<0.5) OR `cog_auc_max_gap_ms` is large (>250ms)), this likely reflects missing data during critical periods
   - These trials should be **excluded** or flagged for further investigation

4. **Use the Diagnostic Scatter Plots**
   - Points in **lower-right quadrant with small gaps** (high quality, low dilation, small gap): Genuine low arousal - **KEEP**
   - Points in **lower-left quadrant or with large gaps** (low quality OR large gap, low dilation): Data quality artifact - **EXCLUDE**
   - **Flat regression slope**: Quality and AUC are independent (good sign - low AUC is physiological)
   - **Positive regression slope**: Suggests quality-AUC relationship (data quality issue)

**Additional Considerations:**

- **Gap-Based Quality** (NOW IMPLEMENTED): We now compute `max_gap_ms` (largest contiguous missing segment) in the cognitive window. Gaps >250ms during the peak response period can severely underestimate AUC even when percent-valid looks acceptable. Following Kret & Sjak-Shie (2019) and Burg et al., we exclude trials with gaps >250ms in primary analyses.
- **Window Duration**: Short windows (fast RTs) can yield less stable AUC estimates. We recommend minimum window duration of 0.5s for primary analyses, with sensitivity checks at 0.3s.
- **Minimum Valid Samples**: At 250 Hz, a 0.5s window should have ~125 samples. With 60% validity, that's ~75 valid samples. We recommend `n_valid >= 100` for primary analyses to ensure robust estimates.
- **Baseline Quality**: Low baseline quality can distort baseline correction, affecting cognitive AUC. Consider excluding trials with `baseline_quality < 0.5`.

**Recommendations:**
- **High Quality + Small Gap + Low RT-Normalized AUC**: Include in analyses (genuine low dilation)
- **Low Quality OR Large Gap + Low RT-Normalized AUC**: Exclude (data quality artifact)
- **Mixed Quality**: Use quality thresholds (50% for Chapter 3, 60% for Chapter 2) + gap-aware exclusions to filter trials

## 9. Conclusion and Next Steps

### 9.1 Data Readiness Summary

**Quality Metrics Implementation:**
- ✅ **B1_quality implemented**: Measures B1 baseline quality (3.85s-4.35s) for Cognitive AUC
- ✅ **Gates properly aligned**: Chapter 2 and Chapter 3 gates use B1_quality (not baseline_quality) for Cognitive AUC analyses
- ✅ **Parallel structure**: 
  - `baseline_quality` (B0): For Total AUC baseline
  - `B1_quality`: For Cognitive AUC baseline
  - `cog_quality`: For Cognitive AUC response window

**Chapter 2 (Psychometric-Pupil Coupling):**
-   **Primary Analysis**: `r format(ch2_primary_total, big.mark = ",")` trials ready (B1 ≥ 60% AND cog ≥ 60%)
-   **Retention Rate**: `r sprintf("%.1f", ch2_rate)`% of total trials
-   **Dataset**: `r format(nrow(ch2_data), big.mark = ",")` trials in `ch2_triallevel.csv` (filter by `gate_pupil_primary == TRUE`)
-   **Gate Definition**: `gate_B1_60` AND `gate_cog_60` AND `gate_auc_both` AND `found_in_flat_run`

**Chapter 3 (DDM with Pupil Predictors):**
-   **Primary Analysis**: `r format(ch3_ready_total, big.mark = ",")` trials ready (B1 ≥ 50% AND cog ≥ 50% + RT filter: 0.2-3.0s)
-   **Retention Rate**: `r sprintf("%.1f", ch3_rate)`% of total trials
-   **Dataset**: `r format(nrow(ch3_data), big.mark = ",")` trials in `ch3_triallevel.csv` (filter by `ddm_ready == TRUE`)
-   **Gate Definition**: `B1_quality >= 0.50` AND `cog_quality >= 0.50` AND RT 0.2-3.0s
-   **Behavior-Only DDM**: `r format(behavior_only_trials, big.mark = ",")` trials available (no pupil quality requirements)

**Overall Data Quality:**
-   **AUC Availability**: `r sprintf("%.1f", auc_available_both)`% of trials have both Total AUC and Cognitive AUC
-   **Behavioral Join Rate**: `r sprintf("%.1f", behavioral_rate)`% of trials have behavioral data
-   **B1_quality Availability**: `r if ("B1_quality" %in% names(merged_data)) sprintf("%.1f", 100 * sum(!is.na(merged_data$B1_quality)) / nrow(merged_data)) else "N/A"`% of trials have B1_quality calculated

### 9.2 Recommended Analyses

1.  **Chapter 2 Primary**: Psychometric coupling with 60% threshold
    (high quality)
2.  **Chapter 2 Sensitivity**: Repeat analyses at 50% and 70% thresholds
3.  **Chapter 3 Primary**: DDM with pupil predictors using 50% threshold
4.  **Chapter 3 Comparison**: Behavior-only DDM for comparison and
    robustness

### 9.3 Data Files

All detailed data files are available in: - `data/pupil_processed/qc/` -
Quality control summaries and gate pass rates -
`data/pupil_processed/analysis_ready/` - Analysis-ready datasets: -
`ch2_triallevel.csv` - Chapter 2 ready data -
`ch3_triallevel.csv` - Chapter 3 ready data -
`data/pupil_processed/merged/` - Full merged trial-level dataset

------------------------------------------------------------------------

**Report Generated**: `r format(Sys.time(), '%B %d, %Y at %I:%M %p')`

**Data Source**: BAP Pupillometry Analysis Pipeline\
**For Questions**: Please refer to the pipeline documentation in
`02_pupillometry_analysis/README.md`
